---
title: "DownstreamAnalysis"
output: html_document
date: "2024-10-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(DESeq2)
library(tidyverse)
library(dplyr)
library(ggrepel)
library(pheatmap)
library(regionReport)
library(ggvenn)
library(VennDiagram)
library(UpSetR)
```

#### What we need:
1. raw count table for DESeq2
2. count table with TPM counts to filter out low expressed genes
3. vst count table for PCA analysis to detect potential outliers

--> before DESeq2 adjust the raw count table with kept genes and samples

### Input data

```{r}
# load count matrix and SRA run table for meta-info
count_matrix_tpm <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/count_matrix.csv', row.names = 1)
count_matrix_tpm <- count_matrix_tpm[, order(colnames(count_matrix_tpm))]
count_matrix_raw <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/count_matrix_raw.csv', row.names = 1)
count_matrix_raw <- count_matrix_raw[, order(colnames(count_matrix_raw))]

sra_runtable <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/SraRunTable-2.txt', row.names = 1)
sra_runtable <- sra_runtable[sra_runtable$treatment == "untreated" | sra_runtable$treatment == "Fusarium oxysporum 5176",]

# meta info with days post treatment, and treatment
meta_info <- sra_runtable
meta_info <- meta_info[rownames(meta_info) != "SRR13824038",]
meta_info$days_post_treatment <- factor(meta_info$days_post_treatment)
meta_info_short <- meta_info[, -c(1:27,30)]
```

### Counting reads (not needed)

```{r}
get_reads_per_sample <- function(counts, type, metadata){
  
  total_reads_per_sample <- colSums(counts)
  total_reads_per_sample_df <- data.frame(sample = names(total_reads_per_sample), total_counts = total_reads_per_sample)

  if(type == "df"){
    return (total_reads_per_sample_df)
  } else if (type == "meta"){
    # merge with meta data
    total_reads_per_sample_meta <- merge(total_reads_per_sample_df, metadata, by.x = "sample", by.y = "Sample_ID")
    return(total_reads_per_sample_meta)
  } else {
    return (total_reads_per_sample)
  }
}

meta_info2 <- meta_info
meta_info2$Sample_ID <- row.names(meta_info2)
total_reads_per_sample_meta <- get_reads_per_sample(count_matrix_raw, "meta", meta_info2)

get_top20_genes <- function(counts, top) {
  
  # total reads per gene
  total_reads_per_gene <- rowSums(counts)
  total_reads_per_gene_df <- data.frame(gene = names(total_reads_per_gene), total_counts = total_reads_per_gene)
  sorted_gene_counts <- sort(total_reads_per_gene, decreasing = TRUE)
  if(top == TRUE) {
    top_20_genes <- head(sorted_gene_counts, 20)
    top_20_genes_df <- data.frame(top_20_genes,gene = names(top_20_genes), counts = top_20_genes)
  
    return (top_20_genes_df)
    
  } else {
    bottom_20_genes <- tail(sorted_gene_counts, 20)
    bottom_20_genes_df <- data.frame(bottom_20_genes,gene = names(bottom_20_genes), counts = bottom_20_genes)
    
    return (bottom_20_genes_df)
  }
  
}

top_20_genes <- get_top20_genes(count_matrix_raw, TRUE)
bottom_20_genes <- get_top20_genes(count_matrix_raw, FALSE)

# visualize counts
ggplot(total_reads_per_sample_meta, aes(sample, total_counts, fill = treatment)) +
  geom_bar(stat="identity", position=position_dodge(width = 5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(top_20_genes, aes(gene, counts)) + geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(bottom_20_genes, aes(gene, counts)) + geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
## DESeq2 (with raw counts)

##### TODO:
1. Construct a DESeqDataSet with count table, colData and design formula

--> design formula: put the variable of interest at the end of the formula and make sure the control level is the first level.

2. Pre-filtering with TPM counts

##### Differential expression analysis
1. Run DESeq and results (build result table with name or contrast)
2. Log fold change shrinkage
3. Order result table by p-values

##### Exploring and exporting results
1. MA-plot to show log2 fold changes
2. Read counts for a single gene (plot counts)

##### Multi-factor design (maybe make a copy of the Data set to rerun the analysis)
~treatment + time + treatment:time

##### Data transformations and visualization (vst)
-blind = TRUE --> compare samples in a manner wholly unbiased by the information about experimental groups 
-blind = FALSE --> better when expecting that many of genes will have large differences in counts which are explainable by the experimental design

##### Data quality assessment (clustering and visualization) in parallel to the differential expression testing (with vsd)
1. Heatmap of count matrix
2. Heatmap of the sample-to-sample distances
3. PCA

##### Interactions
- for time-series experiments: 
1. use interaction and the 
2. likelihood ratio test with a reduced model (which does not contain the interaction terms) which will test whether the condition induces a change in gene expression at any time point after the reference level time point (time 0)


Later:
gene level annotations + 
pathway annotations


### Filter low expressed genes with TPM

Result: 
- before filtering: 47928 genes
- after filtering: 19455 genes

```{r}
smallest_tpm <- 0.5
percentage <- 0.51
num_samples <- ncol(count_matrix_tpm)
# genes to keep
keep <- rowSums(count_matrix_tpm >= smallest_tpm) >= percentage * num_samples

count_matrix_tpm2 <- count_matrix_tpm[keep,]
cts <- count_matrix_raw[keep,]
```


```{r}
# create DESeqDataSet from count matrix
dds <- DESeqDataSetFromMatrix(countData = round(cts),
                              colData = meta_info,
                              design = ~treatment)

# drop samples due to low alignment rate 1.2% - 1.6%
to_drop = c("SRR13824039", "SRR13824011", "SRR13823999", "SRR13823997")
dds <- subset(dds, select = !(colnames(dds) %in% to_drop))
dds <- estimateSizeFactors(dds)
dds <- estimateDispersions(dds)
```

### Data quality assessment

Variance stabilization transformation

```{r}
# get vst counts
vsd_dds <- vst(dds, blind = FALSE)
vsd_dds_mf <- vst(ddsMF, blind = FALSE)
vsd_dds_int <- vst(ddsInt, blind = FALSE)
```

Heatmap of count matrix

```{r}
select <- order(rowMeans(counts(dds, normalized=TRUE)), decreasing = TRUE)
pheatmap(assay(vsd_dds)[select,], cluster_rows=FALSE, show_rownames=FALSE, cluster_cols=FALSE, annotation_col=df)

# sample-to-sample distances
sampleDists <- dist(t(assay(vsd_dds)))
sampleDistMatrix <- as.matrix(sampleDists)

rownames(sampleDistMatrix) <- paste(vsd_dds$treatment, vsd_dds$days_post_treatment, sep="-")
colnames(sampleDistMatrix) <- NULL
pheatmap(sampleDistMatrix, clustering_distance_rows = sampleDists, clustering_distance_cols = sampleDists)
```


```{r}
perform_pca <- function(count){
  # log transform the tpm counts
  log_count_matrix <- log2(count + 1)
  # perform PCA
  pca_result <- prcomp(t(log_count_matrix))
  return (pca_result)
}

visualize_pca <- function(data, pca_result, mark_outliers, color_days){
  if(color_days == TRUE){
    if(mark_outliers == TRUE){
    ggplot(data, aes(x = PC1, y = PC2, color = days_post_treatment, shape = treatment)) +
    geom_point() +
    geom_point(data = outliers, color = "blue") + 
    stat_ellipse(aes(color = treatment), type = "norm") +
    stat_ellipse(data = outliers, color = "blue") + 
    geom_text_repel(data = outliers, aes(label = Sample), vjust = -1, hjust = 1.1, color = "black") +
    xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
    ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
    ggtitle("PCA with data ellipses") +
    theme_minimal()
  } else{
    ggplot(data, aes(x = PC1, y = PC2, color = days_post_treatment, shape = treatment)) +
    geom_point() +
    stat_ellipse(aes(color = treatment), type = "norm") +
    xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
    ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
    ggtitle("PCA with data ellipses") +
    theme_minimal()
  }
  } 
  else{
    if(mark_outliers == TRUE){
      ggplot(data, aes(x = PC1, y = PC2, color = treatment)) +
      geom_point() +
      geom_point(data = outliers, color = "blue") + 
      stat_ellipse(aes(color = treatment), type = "norm", linetype = 2) +
      stat_ellipse(type = "t") +
      stat_ellipse(data = outliers, color = "blue") + 
      geom_text_repel(data = outliers, aes(label = Sample), vjust = -1, hjust = 1.1, color = "black") +
      xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
      ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
      ggtitle("PCA with data ellipses") +
      theme_minimal()
    } else{
      ggplot(data, aes(x = PC1, y = PC2, color = treatment)) +
      geom_point() +
      stat_ellipse(aes(color = treatment), type = "norm", linetype = 2) +
      stat_ellipse(type = "t") +
      xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
      ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
      ggtitle("PCA with data ellipses") +
      theme_minimal()
    }
  }
}
```


```{r}
pca_result_dds <- perform_pca(assay(vsd_dds))
pca_result_mf <- perform_pca(assay(vsd_dds_mf))
pca_result_int <- perform_pca(assay(vsd_dds_int))

# Extract the PCA coordinates
pca_data_dds <- data.frame(Sample = rownames(pca_result_dds$x),
                       PC1 = pca_result_dds$x[, 1], 
                       PC2 = pca_result_dds$x[, 2])

pca_data_mf <- data.frame(Sample = rownames(pca_result_mf$x),
                       PC1 = pca_result_mf$x[, 1], 
                       PC2 = pca_result_mf$x[, 2])

pca_data_int <- data.frame(Sample = rownames(pca_result_int$x),
                       PC1 = pca_result_int$x[, 1], 
                       PC2 = pca_result_int$x[, 2])

# Combine with metadata
pca_data_dds <- merge(pca_data_dds, meta_info2, by.x = "Sample", by.y = "Sample_ID")
pca_data_mf <- merge(pca_data_mf, meta_info2, by.x = "Sample", by.y = "Sample_ID")
pca_data_int <- merge(pca_data_int, meta_info2, by.x = "Sample", by.y = "Sample_ID")

# mark potential outliers
outliers <- pca_data_mf[abs(pca_data_mf$PC1) > 40 | abs(pca_data_mf$PC2) > 30, ]

# visualize PC1 - PC2
visualize_pca(pca_data_dds, mark_outliers = FALSE, color_days = TRUE, pca_result_dds)
visualize_pca(pca_data_mf, mark_outliers = TRUE, color_days = TRUE, pca_result_mf)
visualize_pca(pca_data_int, mark_outliers = TRUE, color_days = TRUE, pca_result_int)
```



```{r}
# DESeq with design ~treatment
dds <- DESeq(dds)
res <- results(dds)

# order p-values
resOrdered <- res[order(res$pvalue),]
```

Number of adjusted p-values less than 0.1: 8767

```{r}
# copy of DESeqDataSet to rerun analysis using multi-factor design
ddsMF <- dds
design(ddsMF) <- formula(~treatment + days_post_treatment)
ddsMF <- DESeq(ddsMF)
resMF <- results(ddsMF)
```

```{r}
# DESeq with Interaction term
ddsInt <- dds
design(ddsInt) <- formula(~treatment + days_post_treatment + treatment:days_post_treatment)
ddsInt <- DESeq(ddsInt)
resInt <- results(ddsInt)

res_int_treatment <- results(ddsInt, contrast = c("treatment", "Fusarium oxysporum 5176", "untreated"))
res_int_time <- results(ddsInt, contrast = c("days_post_treatment", "6", "0"))
```


```{r}
# interaction results for time effect in untreated vs treated
# tests if the time effect is different untreated compared to treated
res_interaction_day1 <- results(ddsInt, name = "treatmentuntreated.days_post_treatment1")
res_interaction_day2 <- results(ddsInt, name = "treatmentuntreated.days_post_treatment2")
res_interaction_day3 <- results(ddsInt, name = "treatmentuntreated.days_post_treatment3")
res_interaction_day4 <- results(ddsInt, name = "treatmentuntreated.days_post_treatment4")
res_interaction_day6 <- results(ddsInt, name = "treatmentuntreated.days_post_treatment6")

# get significant genes out of the results
sig_genes_day1 <- rownames(res_interaction_day1)[which(res_interaction_day1$padj < 0.05 & abs(res_interaction_day1$log2FoldChange) > 1)]
sig_genes_day2 <- rownames(res_interaction_day2)[which(res_interaction_day2$padj < 0.05 & abs(res_interaction_day2$log2FoldChange) > 1)]
sig_genes_day3 <- rownames(res_interaction_day3)[which(res_interaction_day3$padj < 0.05 & abs(res_interaction_day3$log2FoldChange) > 1)]
sig_genes_day4 <- rownames(res_interaction_day4)[which(res_interaction_day4$padj < 0.05 & abs(res_interaction_day4$log2FoldChange) > 1)]
sig_genes_day6 <- rownames(res_interaction_day6)[which(res_interaction_day6$padj < 0.05 & abs(res_interaction_day6$log2FoldChange) > 1)]
```


### Exploring results

```{r}
plotMA(res, ylim=c(-5,5), main="treatment")
plotMA(resMF, ylim=c(-5,5), main="treatment + days_post_treatment")
plotMA(resInt, ylim=c(-5,5), main="treatment + days_post_treatment + treatment:days_post_treatment")

# p-values
hist(res$pvalue)
hist(resMF$pvalue)
hist(resInt$pvalue)
```


```{r}
### Nice 2 have ###
# Likelihood ratio test for interaction effects
# compare second model (mf) with full interaction model (int)
dds_lrt <- DESeq(ddsInt, test = "LRT", reduced = ~treatment + days_post_treatment)
res_lrt <- results(dds_lrt)


# get significant genes from lrt
sig_genes_lrt <- res_lrt[which(res_lrt$padj < 0.05),]
rownames(sig_genes_int)
```

```{r}
# create a venn diagram to show intersections
venn.plot <- venn.diagram(
  x = list(
    "Day 1" = sig_genes_day1, 
    "Day 2" = sig_genes_day2,
    "Day 3" = sig_genes_day3,
    "Day 4" = sig_genes_day4,
    "Day 6" = sig_genes_day6
  ), 
  filename = "sig_genes_venn.png",
  fill = c("red", "blue", "green", "yellow", "orange")
)

# alternative: upsetR (see https://academic.oup.com/bioinformatics/article/33/18/2938/3884387)
gene_lists <- list(
  Day1 = sig_genes_day1, 
  Day2 = sig_genes_day2,
  Day3 = sig_genes_day3,
  Day4 = sig_genes_day4,
  Day6 = sig_genes_day6
)
upset(fromList(gene_lists))
```


```{r}
# Volcano plots for each timepoint
plot_volcano <- function(res, timepoint){
  EnhancedVolcano(res,
    lab = rownames(res),
    title = paste("Volcano plot for Day", timepoint),
    x = 'log2FoldChange',
    y = 'padj')
}

plot_volcano(res_interaction_day1,1)
plot_volcano(res_interaction_day2,2)
plot_volcano(res_interaction_day3,3)
plot_volcano(res_interaction_day4,4)
plot_volcano(res_interaction_day6,6)
```




### Gene level annotation

on Linux cluster:
installed agat with bioconda and created environment "agat".


Step 1: Keep the longest isoforms of transcripts within the gff file
--> agat_sp_keep_longest_isoform.pl -g Araport11_GFF3_genes_transposons.current.gff -o outfile ]


Step 2: Extract Protein Sequences from the processed transcripts
--> agat_sp_extract_sequences.pl -g Araport11_GFF3_longest_transcript.gff -f Arabidopsis_thaliana.TAIR10.cdna.all.fa -p -o Araport11_proteome.fa


Step 3: Load the result proteome fasta file in mercator

