---
title: "DownstreamAnalysis - Praktische Arbeit"
output: html_output
date: "2024-10-04"
author: Luisa Bender
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(DESeq2)
library(tidyverse)
library(dplyr)
library(ggrepel)
library(pheatmap)
library(ggvenn)
library(VennDiagram)
library(UpSetR)
library(seqinr)
library(knitr)
```

#### What we need:
1. raw count table for DESeq2
2. count table with TPM counts to filter out low expressed genes
3. vst count table for PCA analysis to detect potential outliers

--> before DESeq2 adjust the raw count table with kept genes and samples

### Input data

```{r}
# load count matrix and SRA run table for meta-info
count_matrix_tpm <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/count_matrix.csv', row.names = 1)
count_matrix_tpm <- count_matrix_tpm[, order(colnames(count_matrix_tpm))]
count_matrix_raw <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/count_matrix_raw.csv', row.names = 1)
count_matrix_raw <- count_matrix_raw[, order(colnames(count_matrix_raw))]

sra_runtable <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/SraRunTable-2.txt', row.names = 1)
sra_runtable <- sra_runtable[sra_runtable$treatment == "untreated" | sra_runtable$treatment == "Fusarium oxysporum 5176",]

# meta info with days post treatment, and treatment
meta_info <- sra_runtable
meta_info <- meta_info[rownames(meta_info) != "SRR13824038",]
meta_info$days_post_treatment <- factor(meta_info$days_post_treatment)
meta_info_short <- meta_info[, -c(1:27,30)]
```

### Counting reads (not needed)

```{r}
get_reads_per_sample <- function(counts, type, metadata){
  
  total_reads_per_sample <- colSums(counts)
  total_reads_per_sample_df <- data.frame(sample = names(total_reads_per_sample), total_counts = total_reads_per_sample)

  if(type == "df"){
    return (total_reads_per_sample_df)
  } else if (type == "meta"){
    # merge with meta data
    total_reads_per_sample_meta <- merge(total_reads_per_sample_df, metadata, by.x = "sample", by.y = "Sample_ID")
    return(total_reads_per_sample_meta)
  } else {
    return (total_reads_per_sample)
  }
}

meta_info2 <- meta_info
meta_info2$Sample_ID <- row.names(meta_info2)
total_reads_per_sample_meta <- get_reads_per_sample(count_matrix_raw, "meta", meta_info2)

get_top20_genes <- function(counts, top) {
  
  # total reads per gene
  total_reads_per_gene <- rowSums(counts)
  total_reads_per_gene_df <- data.frame(gene = names(total_reads_per_gene), total_counts = total_reads_per_gene)
  sorted_gene_counts <- sort(total_reads_per_gene, decreasing = TRUE)
  if(top == TRUE) {
    top_20_genes <- head(sorted_gene_counts, 20)
    top_20_genes_df <- data.frame(top_20_genes,gene = names(top_20_genes), counts = top_20_genes)
  
    return (top_20_genes_df)
    
  } else {
    bottom_20_genes <- tail(sorted_gene_counts, 20)
    bottom_20_genes_df <- data.frame(bottom_20_genes,gene = names(bottom_20_genes), counts = bottom_20_genes)
    
    return (bottom_20_genes_df)
  }
  
}

top_20_genes <- get_top20_genes(count_matrix_raw, TRUE)
bottom_20_genes <- get_top20_genes(count_matrix_raw, FALSE)

# visualize counts
ggplot(total_reads_per_sample_meta, aes(sample, total_counts, fill = treatment)) +
  geom_bar(stat="identity", position=position_dodge(width = 5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(top_20_genes, aes(gene, counts)) + geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(bottom_20_genes, aes(gene, counts)) + geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
## DESeq2 (with raw counts)

##### TODO:
1. Construct a DESeqDataSet with count table, colData and design formula

--> design formula: put the variable of interest at the end of the formula and make sure the control level is the first level.

2. Pre-filtering with TPM counts

##### Differential expression analysis
1. Run DESeq and results (build result table with name or contrast)
2. Log fold change shrinkage
3. Order result table by p-values

##### Exploring and exporting results
1. MA-plot to show log2 fold changes
2. Read counts for a single gene (plot counts)

##### Multi-factor design (maybe make a copy of the Data set to rerun the analysis)
~treatment + time + treatment:time

##### Data transformations and visualization (vst)
-blind = TRUE --> compare samples in a manner wholly unbiased by the information about experimental groups 
-blind = FALSE --> better when expecting that many of genes will have large differences in counts which are explainable by the experimental design

##### Data quality assessment (clustering and visualization) in parallel to the differential expression testing (with vsd)
1. Heatmap of count matrix
2. Heatmap of the sample-to-sample distances
3. PCA

##### Interactions
- for time-series experiments: 
1. use interaction and the 
2. likelihood ratio test with a reduced model (which does not contain the interaction terms) which will test whether the condition induces a change in gene expression at any time point after the reference level time point (time 0)


Later:
gene level annotations + 
pathway annotations


### Filter low expressed genes with TPM

Result: 
- before filtering: 47928 genes
- after filtering: 19455 genes

```{r}
smallest_tpm <- 0.5
percentage <- 0.51
num_samples <- ncol(count_matrix_tpm)
# genes to keep
keep <- rowSums(count_matrix_tpm >= smallest_tpm) >= percentage * num_samples

count_matrix_tpm2 <- count_matrix_tpm[keep,]
cts <- count_matrix_raw[keep,]
```


```{r}
# create DESeqDataSet from count matrix
dds <- DESeqDataSetFromMatrix(countData = round(cts),
                              colData = meta_info,
                              design = ~treatment)

# drop samples due to low alignment rate 1.2% - 1.6%
to_drop = c("SRR13824039", "SRR13824011", "SRR13823999", "SRR13823997")
dds <- subset(dds, select = !(colnames(dds) %in% to_drop))
dds <- estimateSizeFactors(dds)
dds <- estimateDispersions(dds)

# multi factor design
ddsMF <- dds
design(ddsMF) <- formula(~treatment + days_post_treatment)

# interaction design
ddsInt <- dds
design(ddsInt) <- formula(~treatment + days_post_treatment + treatment:days_post_treatment)
```

### Data quality assessment

Variance stabilization transformation

```{r}
# get vst counts
vsd_dds <- vst(dds, blind = FALSE)
vsd_dds_mf <- vst(ddsMF, blind = FALSE)
vsd_dds_int <- vst(ddsInt, blind = FALSE)
```

Heatmap of count matrix

```{r}
select <- order(rowMeans(counts(dds, normalized=TRUE)), decreasing = TRUE)
pheatmap(assay(vsd_dds)[select,], cluster_rows=FALSE, show_rownames=FALSE, cluster_cols=FALSE, annotation_col=df)

# sample-to-sample distances
sampleDists <- dist(t(assay(vsd_dds)))
sampleDistMatrix <- as.matrix(sampleDists)

rownames(sampleDistMatrix) <- paste(vsd_dds$treatment, vsd_dds$days_post_treatment, sep="-")
colnames(sampleDistMatrix) <- NULL
pheatmap(sampleDistMatrix, clustering_distance_rows = sampleDists, clustering_distance_cols = sampleDists)
```


```{r}
perform_pca <- function(count){
  # log transform the tpm counts
  log_count_matrix <- log2(count + 1)
  # perform PCA
  pca_result <- prcomp(t(log_count_matrix))
  return (pca_result)
}

visualize_pca <- function(data, pca_result, mark_outliers, color_days){
  if(color_days == TRUE){
    if(mark_outliers == TRUE){
    ggplot(data, aes(x = PC1, y = PC2, color = days_post_treatment, shape = treatment)) +
    geom_point() +
    geom_point(data = outliers, color = "blue") + 
    stat_ellipse(aes(color = treatment), type = "norm") +
    stat_ellipse(data = outliers, color = "blue") + 
    geom_text_repel(data = outliers, aes(label = Sample), vjust = -1, hjust = 1.1, color = "black") +
    xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
    ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
    ggtitle("PCA of vst counts") +
    theme_minimal()
  } else{
    ggplot(data, aes(x = PC1, y = PC2, color = days_post_treatment, shape = treatment)) +
    geom_point() +
    stat_ellipse(aes(color = treatment), type = "norm") +
    xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
    ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
    ggtitle("PCA of vst counts") +
    theme_minimal()
  }
  } 
  else{
    if(mark_outliers == TRUE){
      ggplot(data, aes(x = PC1, y = PC2, color = treatment)) +
      geom_point() +
      geom_point(data = outliers, color = "blue") + 
      stat_ellipse(aes(color = treatment), type = "norm", linetype = 2) +
      stat_ellipse(type = "t") +
      stat_ellipse(data = outliers, color = "blue") + 
      geom_text_repel(data = outliers, aes(label = Sample), vjust = -1, hjust = 1.1, color = "black") +
      xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
      ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
      ggtitle("PCA of vst counts") +
      theme_minimal()
    } else{
      ggplot(data, aes(x = PC1, y = PC2, color = treatment)) +
      geom_point() +
      stat_ellipse(aes(color = treatment), type = "norm", linetype = 2) +
      stat_ellipse(type = "t") +
      xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
      ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
      ggtitle("PCA of vst counts") +
      theme_minimal()
    }
  }
}
```


```{r}
pca_result_dds <- perform_pca(assay(vsd_dds))
pca_result_mf <- perform_pca(assay(vsd_dds_mf))
pca_result_int <- perform_pca(assay(vsd_dds_int))

# Extract the PCA coordinates
pca_data_dds <- data.frame(Sample = rownames(pca_result_dds$x),
                       PC1 = pca_result_dds$x[, 1], 
                       PC2 = pca_result_dds$x[, 2])

pca_data_mf <- data.frame(Sample = rownames(pca_result_mf$x),
                       PC1 = pca_result_mf$x[, 1], 
                       PC2 = pca_result_mf$x[, 2])

pca_data_int <- data.frame(Sample = rownames(pca_result_int$x),
                       PC1 = pca_result_int$x[, 1], 
                       PC2 = pca_result_int$x[, 2])

# Combine with metadata
pca_data_dds <- merge(pca_data_dds, meta_info2, by.x = "Sample", by.y = "Sample_ID")
pca_data_mf <- merge(pca_data_mf, meta_info2, by.x = "Sample", by.y = "Sample_ID")
pca_data_int <- merge(pca_data_int, meta_info2, by.x = "Sample", by.y = "Sample_ID")

# mark potential outliers
outliers <- pca_data_mf[abs(pca_data_mf$PC1) > 40 | abs(pca_data_mf$PC2) > 20, ]

# visualize PC1 - PC2
visualize_pca(pca_data_dds, mark_outliers = TRUE, color_days = TRUE, pca_result_dds)
visualize_pca(pca_data_mf, mark_outliers = TRUE, color_days = TRUE, pca_result_mf)
visualize_pca(pca_data_int, mark_outliers = TRUE, color_days = TRUE, pca_result_int)
```



```{r}
# DESeq with design ~treatment
dds <- DESeq(dds)
res <- results(dds)

# order p-values
resOrdered <- res[order(res$pvalue),]
```

Number of adjusted p-values less than 0.1: 8767

```{r}
# rerun analysis using multi-factor design
ddsMF <- DESeq(ddsMF)
resMF <- results(ddsMF)
```

```{r}
# DESeq with Interaction term
ddsInt <- DESeq(ddsInt)
resInt <- results(ddsInt)

res_int_treatment <- results(ddsInt, contrast = c("treatment", "Fusarium oxysporum 5176", "untreated"))
res_int_time <- results(ddsInt, contrast = c("days_post_treatment", "6", "0"))
```


```{r}
# interaction results for time effect in untreated vs treated
# tests if the time effect is different untreated compared to treated
res_interaction_day1 <- results(ddsInt, name = "treatmentuntreated.days_post_treatment1")
res_interaction_day2 <- results(ddsInt, name = "treatmentuntreated.days_post_treatment2")
res_interaction_day3 <- results(ddsInt, name = "treatmentuntreated.days_post_treatment3")
res_interaction_day4 <- results(ddsInt, name = "treatmentuntreated.days_post_treatment4")
res_interaction_day6 <- results(ddsInt, name = "treatmentuntreated.days_post_treatment6")

# get significant genes out of the results
sig_genes_day1 <- rownames(res_interaction_day1)[which(res_interaction_day1$padj < 0.05 & abs(res_interaction_day1$log2FoldChange) > 1)]
sig_genes_day2 <- rownames(res_interaction_day2)[which(res_interaction_day2$padj < 0.05 & abs(res_interaction_day2$log2FoldChange) > 1)]
sig_genes_day3 <- rownames(res_interaction_day3)[which(res_interaction_day3$padj < 0.05 & abs(res_interaction_day3$log2FoldChange) > 1)]
sig_genes_day4 <- rownames(res_interaction_day4)[which(res_interaction_day4$padj < 0.05 & abs(res_interaction_day4$log2FoldChange) > 1)]
sig_genes_day6 <- rownames(res_interaction_day6)[which(res_interaction_day6$padj < 0.05 & abs(res_interaction_day6$log2FoldChange) > 1)]
```


### Exploring results

```{r}
plotMA(res, ylim=c(-5,5), main="treatment")
plotMA(resMF, ylim=c(-5,5), main="treatment + days_post_treatment")
plotMA(resInt, ylim=c(-5,5), main="treatment + days_post_treatment + treatment:days_post_treatment")

# p-values
hist(res$pvalue)
hist(resMF$pvalue)
hist(resInt$pvalue)
```


```{r}
### Nice 2 have ###
# Likelihood ratio test for interaction effects
# compare second model (mf) with full interaction model (int)
dds_lrt <- DESeq(ddsInt, test = "LRT", reduced = ~treatment + days_post_treatment)
res_lrt <- results(dds_lrt)


# get significant genes from lrt
sig_genes_lrt <- res_lrt[which(res_lrt$padj < 0.05),]
#rownames(sig_genes_int)
```

```{r}
# create a venn diagram to show intersections
venn.plot <- venn.diagram(
  x = list(
    "Day 1" = sig_genes_day1, 
    "Day 2" = sig_genes_day2,
    "Day 3" = sig_genes_day3,
    "Day 4" = sig_genes_day4,
    "Day 6" = sig_genes_day6
  ), 
  filename = "plots/sig_genes_venn.png",
  fill = c("red", "blue", "green", "yellow", "orange")
)

# alternative: upsetR (see https://academic.oup.com/bioinformatics/article/33/18/2938/3884387)
gene_lists <- list(
  Day1 = sig_genes_day1, 
  Day2 = sig_genes_day2,
  Day3 = sig_genes_day3,
  Day4 = sig_genes_day4,
  Day6 = sig_genes_day6
)
upset(fromList(gene_lists))
```


```{r}
# Volcano plots for each timepoint
plot_volcano <- function(res, timepoint){
  EnhancedVolcano(res,
    lab = rownames(res),
    title = paste("Volcano plot for Day", timepoint),
    x = 'log2FoldChange',
    y = 'padj')
}
library(EnhancedVolcano)
plot_volcano(res_interaction_day1,1)
plot_volcano(res_interaction_day2,2)
plot_volcano(res_interaction_day3,3)
plot_volcano(res_interaction_day4,4)
plot_volcano(res_interaction_day6,6)
```

### Gene level annotation

Mercator Results: 
Run with A. thaliana transcript file: Arabidopsis_thaliana.TAIR10.cdna.all.fa


```{r}
# read mercator result
mercator_annotations <- read.fasta('/Users/luisabender/Downloads/Job4112.fa')
mercator_result <- read.delim('/Users/luisabender/Downloads/Job4112.results.txt', sep="\t", header = TRUE, stringsAsFactors = FALSE)
mercator_result <- as.data.frame(lapply(mercator_result, function(x) {
    if(is.character(x)) {
        x <- gsub("^'|'$", "", x)
    }
    return(x)
}))
mercator_result$CATEGORY <- sapply(strsplit(mercator_result$NAME, ".", fixed=TRUE), `[`, 1)
```


```{r}
# combine Mercator results with DESeq results
annotated_deseq_results <- function(res, mercator_annos){
  
  # result as dataframe: gene ID as rownames
  res_df <- as.data.frame(res)
  res_df$gene_id <- rownames(res_df)
  mercator_annos$IDENTIFIER <- toupper(mercator_annos$IDENTIFIER)
  merged_results <- merge(res_df, mercator_annos, by.x="gene_id", by.y = "IDENTIFIER", all.x = TRUE)
  return(merged_results)
}


# Count significant genes in each category
analyze_categories <- function(res_annotated, padj_cutoff=0.05, lfc_cutoff=1) {
    sig_genes <- res_annotated[abs(res_annotated$log2FoldChange) > lfc_cutoff & 
                             res_annotated$padj < padj_cutoff, ]
    
    category_counts <- table(sig_genes$CATEGORY)
    category_props <- prop.table(category_counts)
    
    return(list(counts=category_counts, proportions=category_props))
}

# select the Top 30
get_top_annotated_genes <- function(res_annotated, n=30, createtable) {
  # sort by p-value and log fold change
  res_annotated <- res_annotated[order(res_annotated$padj, -abs(res_annotated$log2FoldChange)),]
  # select top n genes
  top_n <- head(res_annotated, n)
  
  if(createtable==TRUE){
    result.table <- data.frame(
      GENE_ID = top_n$gene_id,
      CATEGORY = top_n$CATEGORY,
      FUNCTION = top_n$NAME,
      P_VALUE = top_n$padj,
      LOG_FOLD_CHANGE = top_n$log2FoldChange
    )
    
    return(result.table)
  } else {
    return(top_n)
  }
}

add_timepoint <- function(result, timepoint){
  result <- cbind(
    result,
    days_post_treatment = timepoint
  )
  return(result)
}
```


```{r}
res_day1_annotated <- annotated_deseq_results(res_interaction_day1, mercator_result) %>% write.csv(file="files/res_day1_annotated.csv")
res_day2_annotated <- annotated_deseq_results(res_interaction_day2, mercator_result) %>% write.csv(file="files/res_day2_annotated.csv")
res_day3_annotated <- annotated_deseq_results(res_interaction_day3, mercator_result) %>% write.csv(file="files/res_day3_annotated.csv")
res_day4_annotated <- annotated_deseq_results(res_interaction_day4, mercator_result) %>% write.csv(file="files/res_day4_annotated.csv")
res_day6_annotated <- annotated_deseq_results(res_interaction_day6, mercator_result) %>% write.csv(file="files/res_day6_annotated.csv")

# list of category counts and proportions
cat1 <- analyze_categories(res_day1_annotated)
cat2 <- analyze_categories(res_day2_annotated)
cat3 <- analyze_categories(res_day3_annotated)
cat4 <- analyze_categories(res_day4_annotated)
cat6 <- analyze_categories(res_day6_annotated)
```

```{r}
# Get table for each timepoint
result_day1_top30 <- get_top_annotated_genes(res_day1_annotated, createtable = TRUE) %>% add_timepoint(1) %>% kable(format = "markdown")
result_day2_top30 <- get_top_annotated_genes(res_day2_annotated, createtable = TRUE) %>% add_timepoint(2) %>% kable(format = "markdown")
result_day3_top30 <- get_top_annotated_genes(res_day3_annotated, createtable = TRUE) %>% add_timepoint(3) %>% kable(format = "markdown")
result_day4_top30 <- get_top_annotated_genes(res_day4_annotated, createtable = TRUE) %>% add_timepoint(4) %>% kable(format = "markdown")
result_day6_top30 <- get_top_annotated_genes(res_day6_annotated, createtable = TRUE) %>% add_timepoint(6) %>% kable(format = "markdown")

plot_category_data <- function(category_results, timepoint) {
    df <- data.frame(
        Category = names(category_results$counts),
        Count = as.numeric(category_results$counts),
        Proportion = as.numeric(category_results$proportions),
        Timepoint = timepoint
    )
    return(df)
}

plot_data <- rbind(
    plot_category_data(cat1, "Day1"),
    plot_category_data(cat2, "Day2"),
    plot_category_data(cat3, "Day3"),
    plot_category_data(cat4, "Day4"),
    plot_category_data(cat6, "Day6")

)

ggplot(plot_data, aes(x=Category, y = Count, fill = Timepoint)) +
    geom_bar(stat="identity", position="dodge") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle=90, hjust=1)) +
    labs(title="Distribution of Significant Genes Across Categories",
         x="Functional Category",
         y="Number of Significant Genes")
```


```{r}
plot_data2 <- plot_data[order(-plot_data$Count),]
ggplot(plot_data2, aes(x=Timepoint, y=Category)) + geom_point(aes(size = Count)) + theme_minimal()
```

```{r}
ggplot(result_day1_top30, aes(CATEGORY)) + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Day 1")
ggplot(result_day2_top30, aes(CATEGORY)) + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Day 2")
ggplot(result_day3_top30, aes(CATEGORY)) + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Day 3")
ggplot(result_day4_top30, aes(CATEGORY)) + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Day 4")
ggplot(result_day6_top30, aes(CATEGORY)) + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("Day 6")
```

```{r}
result_day1_top30
result_day2_top30
result_day3_top30
result_day4_top30
result_day6_top30
```


### Pathway level enrichment

Mercator4 BIN enrichment analysis (https://www.plabipd.de/mercator_main.html)

 - One-sided fisher's exact test -> Over-representation analysis
 - genes of interest -> significant genes from each day (sig_genes_dayx)
 - background genes -> all gene IDs from Mercator fasta file (grep -o -E "^>\w+\.\d+" Job4112.fa | tr -d ">" > "geneIDs.txt")


```{r}
# extract genes of interest (per day) for upload into mercator enrichment analysis
write_lines(sig_genes_day1, "files/sig_genes_day1.txt")
write_lines(sig_genes_day2, "files/sig_genes_day2.txt")
write_lines(sig_genes_day3, "files/sig_genes_day3.txt")
write_lines(sig_genes_day4, "files/sig_genes_day4.txt")
write_lines(sig_genes_day6, "files/sig_genes_day6.txt")
```

```{r}
# load MapMan results (with significant pathways)
mapman_day1 <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/Mercator_bin_enrichment_day1.tsv', sep = "\t")
mapman_day2 <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/Mercator_bin_enrichment_day2.tsv', sep = "\t")
mapman_day3 <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/Mercator_bin_enrichment_day3.tsv', sep = "\t")
mapman_day4 <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/Mercator_bin_enrichment_day4.tsv', sep = "\t")
mapman_day6 <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/Mercator_bin_enrichment_day6.tsv', sep = "\t")

# combine all results into one data frame
all_mapman_results <- rbind(
  add_timepoint(mapman_day1,1),
  add_timepoint(mapman_day2,2),
  add_timepoint(mapman_day3,3),
  add_timepoint(mapman_day4,4),
  add_timepoint(mapman_day6,6)
)
all_mapman_results$Category <- sapply(strsplit(all_mapman_results$Context.of.Protein.Function, ".", fixed=TRUE), `[`, 1)
```


```{r}
get_enrichment_of_category <- function(mapman_result, category_number, category){
  mapman_category <- all_mapman_results[startsWith(mapman_result$MapMan4.category.number, category_number),]
  mapman_category <- mapman_category[order(mapman_category$MapMan4.category.number),]
  return(mapman_category)
}

plot_enrichment_of_category <- function(mapman_result, category_number, category){
  data <- get_enrichment_of_category(mapman_result, category_number, category)
  
  ggplot(data, aes(x = days_post_treatment, y = Enrichment.Factor, color = FDR.adjusted.p.value)) +
  geom_point(aes(size = X.Genes.of.Interest.IN.MapMan4.category)) +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  labs(
    x = "Day post Treatment",
    y = "Enrichment Factor",
    color = "FDR adjusted p-value",
    size = "Number of Genes of Interest",
    title = paste("Enrichment factor of genes in", category)) +
  theme_minimal()
}
```

```{r}
plot_enrichment_of_category(all_mapman_results, "21", "Cell wall organisation")
plot_enrichment_of_category(all_mapman_results, "19", "Protein homeostasis")
```

```{r}
# focusing on cell wall
cell_wall_enrichment <- get_enrichment_of_category(all_mapman_results, "21", "Cell wall organisation")
ggplot(cell_wall_enrichment, aes(x=MapMan4.category.number, y = Enrichment.Factor, fill = days_post_treatment)) +
    geom_bar(stat="identity", position="dodge") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle=90, hjust=1)) +
    labs(title=paste("Distribution of Significant Pathways in Cell wall organisation"),
         x="Functional Category",
         y="Enrichment Factor")

ggplot(cell_wall_enrichment, aes(x=MapMan4.category.number, y = Enrichment.Factor)) +
    geom_bar(stat="identity", position="dodge") +
    facet_wrap(vars(days_post_treatment)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle=90, hjust=1)) +
    labs(title="Distribution of Significant Pathways in Cell wall organisation",
         x="Functional Category",
         y="Enrichment Factor")

cell_wall_genes_day2 <- res_day2_annotated[res_day2_annotated$CATEGORY == "Cell wall organisation",]
```

```{r}
# best
ggplot(all_mapman_results, aes(x = days_post_treatment, y = Category)) +
  geom_point(aes(size = -log10(FDR.adjusted.p.value), color = Enrichment.Factor)) +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  labs(
    x = "Timepoint",
    y = "Pathway Category",
    size = "-log10(Adjusted P-Value)",
    color = "Enrichment Factor"
  ) +
  theme_minimal()
```

```{r}
# other visualizations
ggplot(all_mapman_results, aes(x = days_post_treatment, y = Enrichment.Factor, color = Category, group = Category)) +
  geom_line() +
  geom_point(aes(size = X.Genes.of.Interest.IN.MapMan4.category)) +
  labs(
    x = "Day post Treatment",
    y = "Enrichment Factor",
    color = "Pathway Category",
    size = "Number of Genes of Interest"
  ) +
  theme_minimal()

ggplot(all_mapman_results, aes(x = Category, y = Enrichment.Factor, fill = factor(days_post_treatment))) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(
    x = "Pathway Category",
    y = "Enrichment Factor",
    fill = "Timepoint"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```




