---
title: "DownstreamAnalysis"
output: html_document
date: "2024-10-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(DESeq2)
library(tidyverse)
library(dplyr)
library(ggrepel)
library(pheatmap)
```

#### What we need:
1. raw count table for DESeq2
2. count table with TPM counts to filter out low expressed genes
3. vst count table for PCA analysis to detect potential outliers

--> before DESeq2 adjust the raw count table with kept genes and samples

### Input data

```{r}
# load count matrix and SRA run table for meta-info
count_matrix_tpm <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/count_matrix.csv', row.names = 1)
count_matrix_tpm <- count_matrix[, order(colnames(count_matrix))]
count_matrix_raw <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/count_matrix_raw.csv', row.names = 1)
count_matrix_raw <- count_matrix_raw[, order(colnames(count_matrix_raw))]

# exclude 4 samples due to low aligment rate 1.2% - 1.6%

#to_exclude = c("SRR13824039", "SRR13824011", "SRR13823999", "SRR13823997")
#count_matrix_tpm_new <- count_matrix_tpm %>% select(-all_of(to_exclude))
#count_matrix_raw_new <- count_matrix_raw %>% select(-all_of(to_exclude))

sra_runtable <- read.csv('/Users/luisabender/Studium/Praktische Arbeit/SraRunTable-2.txt', row.names = 1)
sra_runtable <- sra_runtable[sra_runtable$treatment == "untreated" | sra_runtable$treatment == "Fusarium oxysporum 5176",]

# meta info with days post treatment, and treatment
#meta_info <- sra_runtable[, -c(1:27,30)]
meta_info <- sra_runtable
meta_info <- meta_info[rownames(meta_info) != "SRR13824038",]
meta_info$days_post_treatment <- factor(meta_info$days_post_treatment)
```

### Counting reads

```{r}
get_reads_per_sample <- function(count_matrix, type, metadata){
  
  total_reads_per_sample <- colSums(count_matrix)
  total_reads_per_sample_df <- data.frame(sample = names(total_reads_per_sample), total_counts = total_reads_per_sample)

  if(type == "df"){
    return (total_reads_per_sample_df)
  } else if (type == "meta"){
    # merge with meta data
    total_reads_per_sample_meta <- merge(total_reads_per_sample_df, metadata, by.x = "sample", by.y = "Sample_ID")
    return(total_reads_per_sample_meta)
  } else {
    return (total_reads_per_sample)
  }
}

meta_info2 <- meta_info
meta_info2$Sample_ID <- row.names(meta_info2)
total_reads_per_sample_meta <- get_reads_per_sample(count_matrix_raw, "meta", meta_info2)

get_top20_genes <- function(count_matrix, top) {
  
  # total reads per gene
  total_reads_per_gene <- rowSums(count_matrix)
  total_reads_per_gene_df <- data.frame(gene = names(total_reads_per_gene), total_counts = total_reads_per_gene)
  sorted_gene_counts <- sort(total_reads_per_gene, decreasing = TRUE)
  if(top == TRUE) {
    top_20_genes <- head(sorted_gene_counts, 20)
    top_20_genes_df <- data.frame(top_20_genes,gene = names(top_20_genes), counts = top_20_genes)
  
    return (top_20_genes_df)
    
  } else {
    bottom_20_genes <- tail(sorted_gene_counts, 20)
    bottom_20_genes_df <- data.frame(bottom_20_genes,gene = names(bottom_20_genes), counts = bottom_20_genes)
    
    return (bottom_20_genes_df)
  }
  
}

top_20_genes <- get_top20_genes(count_matrix_raw_new, TRUE)
bottom_20_genes <- get_top20_genes(count_matrix_raw_new, FALSE)

# visualize counts
ggplot(total_reads_per_sample_meta, aes(sample, total_counts, fill = treatment)) +
  geom_bar(stat="identity", position=position_dodge(width = 5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(top_20_genes, aes(gene, counts)) + geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(bottom_20_genes, aes(gene, counts)) + geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Filter low expressed genes with TPM

Result: 
before filtering: 47928 genes
after filtering: 19455 genes

```{r}
smallest_tpm <- 0.5
percentage <- 0.51
num_samples <- ncol(count_matrix_tpm)
# genes to keep
keep <- rowSums(count_matrix_tpm >= smallest_tpm) >= percentage * num_samples

count_matrix_tpm2 <- count_matrix_tpm[keep,]
cts <- count_matrix_raw[keep,]
```

## DESeq2 (with raw counts)

##### TODO:
1. Construct a DESeqDataSet with count table, colData and design formula
design formula: put the variable of interest at the end of the formula and make sure the control level is the first level.
2. Pre-filtering with TPM counts

Differential expression analysis
1. Run DESeq and results (build result table with name or contrast)
2. Log fold change shrinkage
3. Order result table by p-values

Exploring and exporting results
1. MA-plot to show log2 fold changes
2. Read counts for a single gene (plot counts)

Multi-factor design (maybe make a copy of the Data set to rerun the analysis)
~treatment + time + treatment:time

Data transformations and visualization (vst)
blind = TRUE --> compare samples in a manner wholly unbiased by the information about experimental groups 
blind = FALSE --> better when expecting that many of genes will have large differences in counts which are explainable by the experimental design

Data quality assessment (clustering and visualization) in parallel to the differential expression testing (with vsd)
1. Heatmap of count matrix
2. Heatmap of the sample-to-sample distances
3. PCA

Interactions
plotCounts
for time-series experiments: 1. use interaction and the 
2. likelihood ratio test with a reduced model (which does not contain the interaction terms) which will test whether the condition induces a change in gene expression at any time point after the reference level time point (time 0)

Later: Gene annotation

```{r}
# create DESeqDataSet from count matrix
dds <- DESeqDataSetFromMatrix(countData = round(cts),
                              colData = meta_info,
                              design = ~treatment)

```

```{r}
# DESeq
dds <- DESeq(dds)
res <- results(dds)

# log fold change shrinkage
#resLFC <- lfcShrink(dds, coef="treatment_untreated_vs_Fusarium.oxysporum.5176")
# order p-values
resOrdered <- res[order(res$pvalue),]

```

Number of adjusted p-values less than 0.1: 8767

### Exploring results

```{r}
plotMA(res, ylim=c(-2,2))
# alternativ shrinkage estimators??

# gene with smallest p-value
plotCounts(dds, gene=which.min(res$padj), intgroup="treatment")
```


```{r}
# extract time samples for DESeq2
meta_info_day4 <- meta_info2[meta_info2$days_post_treatment == 4,]
samples_day4 <- meta_info_day4$Sample_ID
count_matrix_day4 <- count_matrix_raw_new %>% select(all_of(samples_day4)) %>% as.matrix()
```


### Run PCA with vst counts

```{r}
perform_pca <- function(count_matrix){
  # log transform the tpm counts
  log_count_matrix <- log2(count_matrix + 1)
  # perform PCA
  pca_result <- prcomp(t(log_count_matrix))
  return (pca_result)
}

visualize_pca <- function(data, pca_result, mark_outliers, color_days){
  if(color_days == TRUE){
    if(mark_outliers == TRUE){
    ggplot(data, aes(x = PC1, y = PC2, color = days_post_treatment, shape = treatment)) +
    geom_point() +
    geom_point(data = outliers, color = "blue") + 
    stat_ellipse(aes(color = treatment), type = "norm") +
    stat_ellipse(data = outliers, color = "blue") + 
    geom_text_repel(data = outliers, aes(label = Sample), vjust = -1, hjust = 1.1, color = "black") +
    xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
    ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
    ggtitle("PCA with data ellipses") +
    theme_minimal()
  } else{
    ggplot(data, aes(x = PC1, y = PC2, days_post_treatment, shape = treatment)) +
    geom_point() +
    stat_ellipse(aes(color = treatment), type = "norm") +
    xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
    ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
    ggtitle("PCA with data ellipses") +
    theme_minimal()
  }
  } else{
    if(mark_outliers == TRUE){
      ggplot(data, aes(x = PC1, y = PC2, color = treatment)) +
      geom_point() +
      geom_point(data = outliers, color = "blue") + 
      stat_ellipse(aes(color = treatment), type = "norm", linetype = 2) +
      stat_ellipse(type = "t") +
      stat_ellipse(data = outliers, color = "blue") + 
      geom_text_repel(data = outliers, aes(label = Sample), vjust = -1, hjust = 1.1, color = "black") +
      xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
      ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
      ggtitle("PCA with data ellipses") +
      theme_minimal()
    } else{
      ggplot(data, aes(x = PC1, y = PC2, color = treatment)) +
      geom_point() +
      stat_ellipse(aes(color = treatment), type = "norm", linetype = 2) +
      stat_ellipse(type = "t") +
      xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
      ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
      ggtitle("PCA with data ellipses") +
      theme_minimal()
    }
  }
}
```


```{r}
# get vst counts
# vst with dds??
count_matrix <- as.matrix(count_matrix_raw)
vsd <- vst(round(count_matrix), blind = FALSE)
#meanSdPlot(assay(vsd))
```


```{r}
pca_result <- perform_pca(as.data.frame(vsd))

# Extract the PCA coordinates
pca_data <- data.frame(Sample = rownames(pca_result$x),
                       PC1 = pca_result$x[, 1], 
                       PC2 = pca_result$x[, 2], 
                       PC3 = pca_result$x[, 3], 
                       PC4 = pca_result$x[, 4])

# Combine with metadata
pca_data <- merge(pca_data, meta_info2, by.x = "Sample", by.y = "Sample_ID")

# mark potential outliers
outliers <- pca_data[abs(pca_data$PC1) > 180 | abs(pca_data$PC2) > 150, ]

# visualize PC1 - PC2
visualize_pca(pca_data,mark_outliers = TRUE, color_days = TRUE, pca_result)
```


```{r}
# include PC3
ggplot(pca_data, aes(x = PC2, y = PC3, color = days_post_treatment, shape = treatment)) +
    geom_point() +
    geom_point(data = outliers, color = "blue") + # mark outliers
    stat_ellipse(aes(color = treatment), type = "norm") +
    stat_ellipse(data = outliers, color = "blue") + 
    geom_text_repel(data = outliers, aes(label = Sample), vjust = -1, hjust = 1.1, color = "black") +
    xlab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
    ylab(paste0("PC3: ", round(100 * summary(pca_result)$importance[2, 3], 2), "% variance")) +
    ggtitle("PCA with data ellipses") +
    theme_minimal()

ggplot(pca_data, aes(x = PC3, y = PC4, color = days_post_treatment, shape = treatment)) +
    geom_point() +
    geom_point(data = outliers, color = "blue") + # mark outliers
    stat_ellipse(data = outliers, color = "blue") + 
    stat_ellipse(aes(color = treatment), type = "norm") +
    geom_text_repel(data = outliers, aes(label = Sample), vjust = -1, hjust = 1.1, color = "black") +
    xlab(paste0("PC3: ", round(100 * summary(pca_result)$importance[2, 3], 2), "% variance")) +
    ylab(paste0("PC4: ", round(100 * summary(pca_result)$importance[2, 4], 2), "% variance")) +
    ggtitle("PCA with data ellipses") +
    theme_minimal()

```





```{r pressure, echo=FALSE}
# for plots without showing the code
plot(pressure)
```


