---
title: "VST and PCA"
output: html_document
date: "2025-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("DESeq2")
```

```{r}
library(ggplot2)
library(DESeq2)
library(dplyr)
library(tidyr)
```

Kallisto alignment result: 

```{r}
kallisto_alignment_summary <- read.csv('~/Studium/bachelor-thesis-luisa/scripts/kallisto_alignment_summary.csv')
poor_alignment_threshold <- 2500000
kallisto_alignment_summary$alignment_status <- ifelse(kallisto_alignment_summary$n_pseudoaligned < poor_alignment_threshold, "Poor", "Good")

bad_alignment <- kallisto_alignment_summary[kallisto_alignment_summary$alignment_status =="Poor",]

kallisto_alignment_summary$n_not_aligned <- kallisto_alignment_summary$n_processed - kallisto_alignment_summary$n_pseudoaligned

data_long <- kallisto_alignment_summary %>%
  select(Sample = sample_ID, Pseudoaligned = n_pseudoaligned, NotAligned = n_not_aligned) %>%
  pivot_longer(cols = c(Pseudoaligned, NotAligned), names_to = "AlignmentStatus", values_to = "Count")

ggplot(data_long, aes(x=Sample, y=Count, fill = AlignmentStatus)) + 
  geom_bar(stat="identity") +
  scale_fill_manual(values = c("Pseudoaligned" = "steelblue", "NotAligned" = "darkred")) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  scale_x_discrete(breaks = data_long$Sample[seq(1, nrow(data_long), by = 10)]) +  # Show every 10th sample
  
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) +
  labs(title = "Kallisto: Alignment Scores for 283 samples",
       x = "",
       y = "Number of aligned Reads",
       fill = "Alignment Status")

#data_long <- data_long[!data_long$Sample %in% to_drop,]
```

Count tables:

```{r count tables}
counts_raw <- read.csv('~/Studium/Bachelor thesis/bachelor-thesis-luisa/count_tables/count_raw_tx.csv', row.names = 1)
counts_raw <- counts_raw[, order(colnames(counts_raw))] # transcripts

counts_tpm <- read.csv('~/Studium/Bachelor thesis/bachelor-thesis-luisa/count_tables/counts_tpm_tx.csv', row.names = 1)
counts_tpm <- counts_tpm[, order(colnames(counts_tpm))] # transcripts
```


```{r count tables}
counts_geneID <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/counts_genelev_corrected.csv', row.names = 1)
duplicates <- c("ERR7738945.1", "ERR7738946.1", "ERR7738947.1")
counts_geneID <- counts_geneID[,!names(counts_geneID) %in% duplicates] # genelevels

counts_genelev_tpm <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/counts_tpm_abundance2.csv', row.names = 1)
counts_genelev_tpm <- counts_genelev_tpm[,!names(counts_genelev_tpm) %in% duplicates] # genelevel tpm

counts_lengthscaledTPM <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/counts_lengthscaledTPM.csv', row.names = 1) # counts scaled with lenghtscaledTPM
```

Meta info:

```{r meta info}
meta_info <- read.csv('~/Studium/bachelor-thesis-luisa/data/meta_info_all.csv', sep = ';')

# with infected/control
metaData <- read.csv('~/Studium/bachelor-thesis-luisa/data/metaData.csv', sep = ';')
metaData <- merge(metaData, kallisto_alignment_summary, by.x = "Run", by.y = "sample_ID", all.x = TRUE)
```


```{r meta info}
# removing samples with bad alignment rate
experiment_to_drop <- 'Hyaloperonospora arabidopsidis 2'
to_drop <- meta_info[meta_info$Experiment == experiment_to_drop, ]$Run
hpa2 <- c("SRR22565219","SRR22565220","SRR22565221","SRR22565222","SRR22565223","SRR22565224","SRR22565225","SRR22565226","SRR22565227","SRR22565228","SRR22565229","SRR22565230","SRR22565231","SRR22565232","SRR22565233","SRR22565234","SRR22565235","SRR22565236")

#meta_info <- meta_info[!meta_info$Run %in% to_drop,]
#meta_info_general <- meta_info_general[!meta_info_general$Run %in% to_drop,]
counts_geneID <- counts_geneID[, !names(counts_geneID) %in% hpa2]
counts_genelev_tpm <- counts_genelev_tpm[, !names(counts_genelev_tpm) %in% hpa2]
```

Filter low expressed genes (with tpm values (genelevel tpm)): 

Result: 
- before filtering: 27321 genes
- after filtering: 24991 genes

if tiniest dataset with contrast, is 3 samples, then
--> keep genes that have at least 0.5 TPM per sample in at least 3 samples.

```{r}
smallest_tpm <- 0.5
percentage <- 0.51
num_samples <- ncol(counts_genelev_tpm)
# genes to keep
keep <- rowSums(counts_genelev_tpm >= smallest_tpm) >= 3

counts_genelev_filtered <- counts_geneID[keep,]
```


VST on filtered gene count table:

```{r}
#counts_genelev_filtered <- counts_genelev_filtered[, !names(counts_genelev_filtered) %in% hpa2]
dds_genelev_filtered <- DESeqDataSetFromMatrix(countData = round(counts_genelev_filtered),
                              colData = metaData,
                              design = ~Experiment + infected.control)

#dds_genelev_filtered <- DESeq(dds_genelev_filtered)
#dds_filtered <- estimateDispersions(dds_filtered)
vsd_genelev_filtered <- varianceStabilizingTransformation(dds_genelev_filtered, blind = FALSE)
#vsd_genelev_filtered_data <- getVarianceStabilizedData(dds_genelev_filtered)
```


PCA function:

```{r pca function}
perform_pca <- function(count){
   #log_count_matrix <- log2(count + 1)
  # perform PCA
  pca_result <- prcomp(t(count))
  return (pca_result)
}

visualize_standard_pca <- function(data, pca_result, ellipse, title, PC3){
  if(ellipse == TRUE){
    if(PC3 == TRUE){
      ggplot(data, aes(x = PC3, y = PC4, color = Experiment)) +
      geom_point() +
      stat_ellipse(aes(color = Experiment), type = "norm") +
      xlab(paste0("PC3: ", round(100 * summary(pca_result)$importance[2, 3], 2), "% variance")) +
      ylab(paste0("PC4: ", round(100 * summary(pca_result)$importance[2, 4], 2), "% variance")) +
      ggtitle(title) +
      theme_minimal()
    }
    else{
    ggplot(data, aes(x = PC1, y = PC2, color = Experiment)) +
      geom_point() +
      stat_ellipse(aes(color = Experiment), type = "norm") +
      xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
      ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
      ggtitle(title) +
      theme_minimal()
      }
  }
  
  else{
    if(PC3 == TRUE){
      ggplot(data, aes(x = PC3, y = PC4, color = Experiment)) +
      geom_point() +
      xlab(paste0("PC3: ", round(100 * summary(pca_result)$importance[2, 3], 2), "% variance")) +
      ylab(paste0("PC4: ", round(100 * summary(pca_result)$importance[2, 4], 2), "% variance")) +
      ggtitle(title) +
      theme_minimal()
    }
    else{
      ggplot(data, aes(x = PC1, y = PC2, color = Experiment)) +
        geom_point() +
        xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
        ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
        ggtitle(title) +
        theme_minimal()
      }
  }
}
```

PCA on raw (and filtered) vst counts:  
  
```{r pca}
# vst counts filtered with tpm value
pca_filtered <- perform_pca(assay(vsd_genelev_filtered))
pca_data_filtered <- data.frame(Sample = rownames(pca_filtered$x),
                       PC1 = pca_filtered$x[, 1], 
                       PC2 = pca_filtered$x[, 2], 
                       PC3 = pca_filtered$x[, 3], 
                       PC4 = pca_filtered$x[, 4]) %>% merge(metaData, by.x = "Sample", by.y = "Run")
```


PCA on unfiltered vs filtered counts

```{r pca}
# PC1 and PC2
visualize_standard_pca(pca_data_filtered,pca_filtered, FALSE,"PCA of filtered vst counts", FALSE)

# PC3 and PC4
visualize_standard_pca(pca_data_filtered,pca_filtered, FALSE,"PCA of filtered vst counts", TRUE)
```

PCA on further information:

```{r pca}
# number of aligned reads per sample (absolute counts) to include in the PCA
ggplot(pca_data_filtered, aes(x = PC1, y = PC2, color = n_pseudoaligned, size = n_pseudoaligned)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  #stat_ellipse(aes(color = Experiment), type = "norm") +# Color gradient for aligned reads
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "Filtered PCA Plot with Aligned Reads",
    x = "PC 1",
    y = "PC 2",
    size = "Aligned Reads",
    color = "Aligned Reads"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )


#PC3 and PC4
ggplot(pca_data_filtered, aes(x = PC3, y = PC4, color = n_pseudoaligned, size = n_pseudoaligned)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  #stat_ellipse(aes(color = Experiment), type = "norm") +# Color gradient for aligned reads
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "Filtered PCA Plot with Aligned Reads",
    x = "PC 3",
    y = "PC 4",
    size = "Aligned Reads",
    color = "Aligned Reads"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )

# include single and paired in the PCA
ggplot(pca_data_filtered, aes(x = PC1, y = PC2, color = single.paired)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA with filtered vst counts",
    x = "PC 1",
    y = "PC 2",
    color = "single-end/paired-end"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )

ggplot(pca_data_filtered, aes(x = PC3, y = PC4, color = single.paired)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA with filtered vst counts",
    x = "PC 3",
    y = "PC 4",
    color = "single-end/paired-end"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )
```

save expression data (filtered and vst)

```{r}
write.csv(assay(vsd_genelev_filtered), "~/Studium/bachelor-thesis-luisa/count_tables/vst_expression_data.csv")
```


