---
title: "VST and PCA"
output: html_document
date: "2025-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r cars}
summary(cars)
```

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("DESeq2")
```

```{r}
library(ggplot2)
library(DESeq2)
library(dplyr)
library(tidyr)
```

Kallisto alignment result: 

```{r}
kallisto_alignment_summary <- read.csv('~/Studium/Bachelor thesis/kallisto_alignment_summary.csv')
poor_alignment_threshold <- 2500000
kallisto_alignment_summary$alignment_status <- ifelse(kallisto_alignment_summary$n_pseudoaligned < poor_alignment_threshold, "Poor", "Good")

bad_alignment <- kallisto_alignment_summary[kallisto_alignment_summary$alignment_status =="Poor",]

kallisto_alignment_summary$n_not_aligned <- kallisto_alignment_summary$n_processed - kallisto_alignment_summary$n_pseudoaligned

data_long <- kallisto_alignment_summary %>%
  select(Sample = sample_ID, Pseudoaligned = n_pseudoaligned, NotAligned = n_not_aligned) %>%
  pivot_longer(cols = c(Pseudoaligned, NotAligned), names_to = "AlignmentStatus", values_to = "Count")

ggplot(data_long, aes(x=Sample, y=Count, fill = AlignmentStatus)) + 
  geom_bar(stat="identity") +
  scale_fill_manual(values = c("Pseudoaligned" = "steelblue", "NotAligned" = "darkred")) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  scale_x_discrete(breaks = data_long$Sample[seq(1, nrow(data_long), by = 10)]) +  # Show every 10th sample
  
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) +
  labs(title = "Kallisto: Alignment Scores for 283 samples",
       x = "",
       y = "Number of aligned Reads",
       fill = "Alignment Status")

#data_long <- data_long[!data_long$Sample %in% to_drop,]
```

Count tables:

```{r count tables}
counts_raw <- read.csv('~/Studium/Bachelor thesis/bachelor-thesis-luisa/count_tables/count_raw_tx.csv', row.names = 1)
counts_raw <- counts_raw[, order(colnames(counts_raw))] # transcripts

counts_tpm <- read.csv('~/Studium/Bachelor thesis/bachelor-thesis-luisa/count_tables/counts_tpm_tx.csv', row.names = 1)
counts_tpm <- counts_tpm[, order(colnames(counts_tpm))] # transcripts

counts_geneID <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/counts_genelevel.csv', row.names = 1)
duplicates <- c("ERR7738945.1", "ERR7738946.1", "ERR7738947.1")
counts_geneID <- counts_geneID[,!names(counts_geneID) %in% duplicates] # genelevels

counts_genelev_tpm <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/counts_tpm_abundance.csv', row.names = 1)
counts_genelev_tpm <- counts_genelev_tpm[,!names(counts_genelev_tpm) %in% duplicates] # genelevel tpm

counts_lengthscaledTPM <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/counts_lengthscaledTPM.csv', row.names = 1) # counts scaled with lenghtscaledTPM
```

Meta info:

```{r meta info}
meta_info <- read.csv('~/Studium/bachelor-thesis-luisa/data/meta_info_all.csv', sep = ';')
# with infected/control table
meta_info_general <- read.csv('~/Studium/bachelor-thesis-luisa/data/meta_info_all_general.csv', sep = ';')

meta_info_old <- read.csv('')
# removing samples with bad alignment rate
experiment_to_drop <- 'Hyaloperonospora arabidopsidis 2'
to_drop <- meta_info[meta_info$Experiment == experiment_to_drop, ]$Run
hpa2 <- c("SRR22565219","SRR22565220","SRR22565221","SRR22565222","SRR22565223","SRR22565224","SRR22565225","SRR22565226","SRR22565227","SRR22565228","SRR22565229","SRR22565230","SRR22565231","SRR22565232","SRR22565233","SRR22565234","SRR22565235","SRR22565236")
# drop in counts table and meta info
#counts_raw <- counts_raw[, !names(counts_raw) %in% to_drop]
#counts_tpm <- counts_tpm[, !names(counts_tpm) %in% to_drop]
meta_info <- meta_info[!meta_info$Run %in% to_drop,]
meta_info_general <- meta_info_general[!meta_info_general$Run %in% to_drop,]
counts_geneID <- counts_geneID[, !names(counts_geneID) %in% hpa2]
counts_genelev_tpm <- counts_genelev_tpm[, !names(counts_genelev_tpm) %in% hpa2]
# include number of aligned reads into meta_info table
#meta_info <- merge(meta_info, kallisto_alignment_summary, by.x = "Run", by.y = "sample_ID", all.x = TRUE)
#meta_info_general <- merge(meta_info_general, kallisto_alignment_summary, by.x = "Run", by.y = "sample_ID", all.x = TRUE)

```

Filter low expressed genes (with tpm values (genelevel tpm)): 

Result: 
- before filtering: 27321 genes
- after filtering: 19017 genes

--> keep genes that have at least 0.5 TPM per sample in at least 51% or 60% of your samples

```{r}
smallest_tpm <- 0.5
percentage <- 0.51
num_samples <- ncol(counts_genelev_tpm)
# genes to keep
keep <- rowSums(counts_genelev_tpm >= smallest_tpm) >= percentage * num_samples

counts_tpm_filtered <- counts_tpm[keep,]
counts_raw_filtered <- counts_raw[keep,]
counts_genelev_filtered <- counts_geneID[keep,]
```

Perform vst on raw counts (transcripts):

```{r vst}
dds <- DESeqDataSetFromMatrix(countData = round(counts_raw),
                              colData = meta_info_general,
                              design = ~Experiment + infected.control)
#dds <- DESeq(dds)
vsd <- varianceStabilizingTransformation(dds) # blind = FALSE!!
vsd_data <- getVarianceStabilizedData(dds)
```

vst with filtered dataset (transcripts):

```{r vst}
dds_filtered <- DESeqDataSetFromMatrix(countData = round(counts_raw_filtered),
                              colData = meta_info_general,
                              design = ~Experiment + infected.control)
#dds_filtered <- DESeq(dds_filtered)
#dds_filtered <- estimateDispersions(dds_filtered)
vsd_filt <- varianceStabilizingTransformation(dds_filtered, blind = FALSE)
#vsd_filt_data <- getVarianceStabilizedData(dds_filtered)
```

vst on gene count table: 

```{r}
dds_genes <- DESeqDataSetFromMatrix(countData = round(counts_geneID),
                              colData = meta_info_general,
                              design = ~Experiment + infected.control)
#dds_filtered <- DESeq(dds_filtered)
#dds_filtered <- estimateDispersions(dds_filtered)
vsd_genes <- varianceStabilizingTransformation(dds_genes, blind = FALSE)
```

without outlier:

```{r}
dds_genes2 <- DESeqDataSetFromMatrix(countData = round(counts_geneID),
                              colData = meta_info_general,
                              design = ~Experiment + infected.control)
#dds_filtered <- DESeq(dds_filtered)
#dds_filtered <- estimateDispersions(dds_filtered)
vsd_genes2 <- varianceStabilizingTransformation(dds_genes2, blind = FALSE)
```

filtered count table:

```{r}
dds_genelev_filtered <- DESeqDataSetFromMatrix(countData = round(counts_genelev_filtered),
                              colData = meta_info_general,
                              design = ~Experiment + infected.control)
#dds_filtered <- DESeq(dds_filtered)
#dds_filtered <- estimateDispersions(dds_filtered)
vsd_genelev_filtered <- varianceStabilizingTransformation(dds_genelev_filtered, blind = FALSE)
```

vst on tpm counts (tpm values):

```{r}
dds_tpm <- DESeqDataSetFromMatrix(countData = round(counts_genelev_tpm),
                              colData = meta_info_general,
                              design = ~Experiment + infected.control)
#dds_filtered <- DESeq(dds_filtered)
#dds_filtered <- estimateDispersions(dds_filtered)
vsd_tpm <- varianceStabilizingTransformation(dds_tpm, blind = FALSE)
```

PCA function:

```{r pca function}
perform_pca <- function(count){
   #log_count_matrix <- log2(count + 1)
  # perform PCA
  pca_result <- prcomp(t(count))
  return (pca_result)
}

visualize_standard_pca <- function(data, pca_result, ellipse, title, PC3){
  if(ellipse == TRUE){
    if(PC3 == TRUE){
      ggplot(data, aes(x = PC3, y = PC4, color = Experiment)) +
      geom_point() +
      stat_ellipse(aes(color = Experiment), type = "norm") +
      xlab(paste0("PC3: ", round(100 * summary(pca_result)$importance[2, 3], 2), "% variance")) +
      ylab(paste0("PC4: ", round(100 * summary(pca_result)$importance[2, 4], 2), "% variance")) +
      ggtitle(title) +
      theme_minimal()
    }
    else{
    ggplot(data, aes(x = PC1, y = PC2, color = Experiment)) +
      geom_point() +
      stat_ellipse(aes(color = Experiment), type = "norm") +
      xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
      ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
      ggtitle(title) +
      theme_minimal()
      }
  }
  
  else{
    if(PC3 == TRUE){
      ggplot(data, aes(x = PC3, y = PC4, color = Experiment)) +
      geom_point() +
      xlab(paste0("PC3: ", round(100 * summary(pca_result)$importance[2, 3], 2), "% variance")) +
      ylab(paste0("PC4: ", round(100 * summary(pca_result)$importance[2, 4], 2), "% variance")) +
      ggtitle(title) +
      theme_minimal()
    }
    else{
      ggplot(data, aes(x = PC1, y = PC2, color = Experiment)) +
        geom_point() +
        xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
        ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
        ggtitle(title) +
        theme_minimal()
      }
  }
}
```

PCA on raw (and filtered) vst counts:  
  
```{r pca}
# vst counts filtered with tpm value
pca_filtered <- perform_pca(assay(vsd_genelev_filtered))
pca_data_filtered <- data.frame(Sample = rownames(pca_filtered$x),
                       PC1 = pca_filtered$x[, 1], 
                       PC2 = pca_filtered$x[, 2], 
                       PC3 = pca_filtered$x[, 3], 
                       PC4 = pca_filtered$x[, 4]) %>% merge(meta_info_general, by.x = "Sample", by.y = "Run")
# contains all datasets
pca_genes <- perform_pca(assay(vsd_genes))
pca_data_genes <- data.frame(Sample = rownames(pca_genes$x),
                       PC1 = pca_genes$x[, 1], 
                       PC2 = pca_genes$x[, 2], 
                       PC3 = pca_genes$x[, 3], 
                       PC4 = pca_genes$x[, 4]) %>% merge(meta_info_general, by.x = "Sample", by.y = "Run")

# Hpa2 filtered out
pca_genes2 <- perform_pca(assay(vsd_genes2))
pca_data_genes2 <- data.frame(Sample = rownames(pca_genes2$x),
                       PC1 = pca_genes2$x[, 1], 
                       PC2 = pca_genes2$x[, 2], 
                       PC3 = pca_genes2$x[, 3], 
                       PC4 = pca_genes2$x[, 4]) %>% merge(meta_info_general, by.x = "Sample", by.y = "Run")
```

PCA on tpm counts (lengthscaledTPM):

```{r}
pca_tpm <- perform_pca(counts_lengthscaledTPM)
pca_data_tpm <- data.frame(Sample = rownames(pca_tpm$x),
                       PC1 = pca_tpm$x[, 1], 
                       PC2 = pca_tpm$x[, 2], 
                       PC3 = pca_tpm$x[, 3], 
                       PC4 = pca_tpm$x[, 4]) %>% merge(meta_info_general, by.x = "Sample", by.y = "Run")

```

```{r}
visualize_standard_pca(pca_data_tpm, pca_tpm, FALSE, "PCA of tpm counts (lengthscaledTPM)", FALSE)
visualize_standard_pca(pca_data_tpm, pca_tpm, FALSE,"PCA of tpm counts (lengthscaledTPM)", TRUE)
```

PCA on raw counts (unchanged):

```{r}
pca_raw <- perform_pca(counts_geneID)
pca_data_raw <- data.frame(Sample = rownames(pca_raw$x),
                       PC1 = pca_raw$x[, 1], 
                       PC2 = pca_raw$x[, 2], 
                       PC3 = pca_raw$x[, 3], 
                       PC4 = pca_raw$x[, 4]) %>% merge(meta_info, by.x = "Sample", by.y = "Run")

visualize_standard_pca(pca_data_raw,pca_raw,TRUE, "PCA of raw counts", FALSE)
visualize_standard_pca(pca_data_raw,pca_raw,TRUE, "PCA of raw counts", TRUE)
```


PCA on unfiltered vs filtered counts

```{r pca}
# PC1 and PC2
visualize_standard_pca(pca_data_genes2,pca_genes2, FALSE,"PCA of vst counts", FALSE)
visualize_standard_pca(pca_data_filtered,pca_filtered, FALSE,"PCA of filtered vst counts", FALSE)
```


```{r pca}
# PC3 and PC4
visualize_standard_pca(pca_data_genes2,pca_genes2, FALSE,"PCA of vst counts", TRUE)
visualize_standard_pca(pca_data_filtered,pca_filtered, FALSE,"PCA of filtered vst counts", TRUE)
ggplot(pca_data_filtered, aes(x = PC1, y = PC2, color = Experiment, shape = infected.control)) +
        geom_point() +
        xlab(paste0("PC1: ", round(100 * summary(pca_filtered)$importance[2, 1], 2), "% variance")) +
        ylab(paste0("PC2: ", round(100 * summary(pca_filtered)$importance[2, 2], 2), "% variance")) +
        ggtitle(title) +
        theme_minimal()
```


PCA on further information:

```{r pca}
# number of aligned reads per sample (absolute counts) to include in the PCA
ggplot(pca_data_genes2, aes(x = PC1, y = PC2, color = n_pseudoaligned, size = n_pseudoaligned)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  #stat_ellipse(aes(color = Experiment), type = "norm") +# Color gradient for aligned reads
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "PCA Plot with Aligned Reads",
    x = "PC 1",
    y = "PC 2",
    size = "Aligned Reads",
    color = "Aligned Reads"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )
ggplot(pca_data_filtered, aes(x = PC1, y = PC2, color = n_pseudoaligned, size = n_pseudoaligned)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  #stat_ellipse(aes(color = Experiment), type = "norm") +# Color gradient for aligned reads
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "Filtered PCA Plot with Aligned Reads",
    x = "PC 1",
    y = "PC 2",
    size = "Aligned Reads",
    color = "Aligned Reads"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )
ggplot(pca_data_raw, aes(x = PC1, y = PC2, color = n_pseudoaligned, size = n_pseudoaligned)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  #stat_ellipse(aes(color = Experiment), type = "norm") +# Color gradient for aligned reads
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "PCA Plot with Aligned Reads (raw counts)",
    x = "PC 1",
    y = "PC 2",
    size = "Aligned Reads",
    color = "Aligned Reads"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )
ggplot(pca_data_tpm, aes(x = PC1, y = PC2, color = n_pseudoaligned, size = n_pseudoaligned)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  #stat_ellipse(aes(color = Experiment), type = "norm") +# Color gradient for aligned reads
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "PCA Plot with Aligned Reads (tpm counts)",
    x = "PC 1",
    y = "PC 2",
    size = "Aligned Reads",
    color = "Aligned Reads"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )

#PC3 and PC4
ggplot(pca_data_genes2, aes(x = PC3, y = PC4, color = n_pseudoaligned, size = n_pseudoaligned)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  #stat_ellipse(aes(color = Experiment), type = "norm") +# Color gradient for aligned reads
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "PCA Plot with Aligned Reads",
    x = "PC 3",
    y = "PC 4",
    size = "Aligned Reads",
    color = "Aligned Reads"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )
ggplot(pca_data_filtered, aes(x = PC3, y = PC4, color = n_pseudoaligned, size = n_pseudoaligned)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  #stat_ellipse(aes(color = Experiment), type = "norm") +# Color gradient for aligned reads
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "Filtered PCA Plot with Aligned Reads",
    x = "PC 3",
    y = "PC 4",
    size = "Aligned Reads",
    color = "Aligned Reads"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )

# include single and paired in the PCA
ggplot(pca_data_genes2, aes(x = PC1, y = PC2, color = single.paired)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA with vst counts",
    x = "PC 1",
    y = "PC 2",
    color = "single-end/paired-end"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )

# PC 3 and PC 4
ggplot(pca_data_genes2, aes(x = PC3, y = PC4, color = single.paired)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA with vst counts",
    x = "PC 3",
    y = "PC 4",
    color = "single-end/paired-end"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )

#filtered
ggplot(pca_data_filtered, aes(x = PC1, y = PC2, color = single.paired)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA with filtered vst counts",
    x = "PC 1",
    y = "PC 2",
    color = "single-end/paired-end"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )

ggplot(pca_data_filtered, aes(x = PC3, y = PC4, color = single.paired)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA with filtered vst counts",
    x = "PC 3",
    y = "PC 4",
    color = "single-end/paired-end"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )
```

save expression data (filtered and vst)

```{r}
write.csv(assay(vsd_genelev_filtered), "~/Studium/bachelor-thesis-luisa/count_tables/vst_expression_data.csv")
```


