---
title: "Data cleaning & preprocessing"
output: html_document
date: "2025-01-27"
---
```{r}
library(WGCNA)
library(ggplot2)
library(DESeq2)
library(dplyr)
library(tidyr)
```

### Kallisto alignment result: 

```{r}
kallisto_alignment_summary <- read.csv('~/Studium/bachelor-thesis-luisa/multiqc/kallisto_alignment_summary.csv')

kallisto_alignment_summary$n_not_aligned <- kallisto_alignment_summary$n_processed - kallisto_alignment_summary$n_pseudoaligned

data_long <- kallisto_alignment_summary %>%
  select(Sample = sample_ID, Pseudoaligned = n_pseudoaligned, NotAligned = n_not_aligned) %>%
  pivot_longer(cols = c(Pseudoaligned, NotAligned), names_to = "AlignmentStatus", values_to = "Count")

ggplot(data_long, aes(x=Sample, y=Count, fill = AlignmentStatus)) + 
  geom_bar(stat="identity") +
  scale_fill_manual(values = c("Pseudoaligned" = "steelblue", "NotAligned" = "darkred")) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  #scale_x_discrete(breaks = data_long$Sample[seq(1, nrow(data_long), by = 10)]) +  # Show every 10th sample
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) +
  labs(title = "Kallisto: Alignment Scores for added dataset (Ecr)",
       x = data_long$Sample,
       y = "Number of aligned Reads",
       fill = "Alignment Status")

#data_long <- data_long[!data_long$Sample %in% to_drop,]
```

Loading count data (saved via tximport on cluster):

```{r count tables}
counts_geneID <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/counts_genelev_corrected.csv', row.names = 1)
duplicates <- c("ERR7738945.1", "ERR7738946.1", "ERR7738947.1")
counts_geneID <- counts_geneID[,!names(counts_geneID) %in% duplicates] # genelevel counts

counts_genelev_tpm <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/counts_tpm_abundance2.csv', row.names = 1)
counts_genelev_tpm <- counts_genelev_tpm[,!names(counts_genelev_tpm) %in% duplicates] # genelevel tpm

counts_lengthscaledTPM <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/counts_lengthscaledTPM.csv', row.names = 1) # counts scaled with lenghtscaledTPM

gene_counts_stegmann <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/RNAseq.gene.counts.kallisto.txt',sep="\t", row.names = 1)
tpm_counts_stegmann <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/RNAseq.gene.abundance.kallisto.txt',sep="\t", row.names = 1)

gene_counts_all <- read.csv('~/Studium/bachelor-thesis-luisa/scripts/gene_counts_w_added_dataset.csv', row.names = 1)
gene_counts_all <- gene_counts_all[, !names(gene_counts_all) %in% duplicates]
abundance_counts_all <- read.csv('~/Studium/bachelor-thesis-luisa/scripts/abundance_counts_w_added_dataset.csv', row.names = 1)
abundance_counts_all <- abundance_counts_all[, !names(abundance_counts_all) %in% duplicates]

rownames(gene_counts_all) <- gsub("\\.1$", "", rownames(gene_counts_all)) 
rownames(abundance_counts_all) <- gsub("\\.1$", "", rownames(abundance_counts_all))
colnames(gene_counts_all) <- gsub('X','',colnames(gene_counts_all)) %>% gsub("_R1_001", "", .)
colnames(abundance_counts_all) <- gsub('X','',colnames(abundance_counts_all)) %>% gsub("_R1_001", "", .)
```

```{r}
# keep only Col0 samples of dataset
stegmann_samples_to_keep <- c("1_5_S4","1_9_S8","2_9_S21","2_13_S25","3_1_S29","3_9_S30","4_1_S31","4_5_S35","4_9_S39","4_13_S43","5_1_S47","5_5_S51","5_9_S55","5_13_S59","6_1_S63","6_5_S67","6_13_S74")

# remove X at the beginning of the sample names
colnames(gene_counts_stegmann) <- gsub('X','',colnames(gene_counts_stegmann))
colnames(tpm_counts_stegmann) <- gsub('X','',colnames(tpm_counts_stegmann))

gene_counts_stegmann <- gene_counts_stegmann[, names(gene_counts_stegmann) %in% stegmann_samples_to_keep]
tpm_counts_stegmann <- tpm_counts_stegmann[, names(tpm_counts_stegmann) %in% stegmann_samples_to_keep]
```


```{r}
# set gene ids as separate column
gene_counts_stegmann <- gene_counts_stegmann %>% tibble::rownames_to_column(var = "gene_id")
counts_geneID <- counts_geneID %>% tibble::rownames_to_column(var = "gene_id")

tpm_counts_stegmann <- tpm_counts_stegmann %>% tibble::rownames_to_column(var = "gene_id")
counts_genelev_tpm <- counts_genelev_tpm %>% tibble::rownames_to_column(var = "gene_id")

counts_geneID$gene_id <- gsub("\\.1$", "", counts_geneID$gene_id)
counts_genelev_tpm$gene_id <- gsub("\\.1$", "", counts_genelev_tpm$gene_id)

# combine to count tables
gene_counts <- full_join(counts_geneID,gene_counts_stegmann, by="gene_id")
tpm_counts <-  full_join(counts_genelev_tpm,tpm_counts_stegmann, by="gene_id")

gene_counts[is.na(gene_counts)] <- 0
tpm_counts[is.na(tpm_counts)] <- 0

# set genes back as rownames
gene_counts <- gene_counts %>% tibble::column_to_rownames(var = "gene_id")
tpm_counts <- tpm_counts %>% tibble::column_to_rownames(var = "gene_id")
```


Loading meta data:

```{r}
metaData <- read.csv("~/Studium/bachelor-thesis-luisa/data/metaData_2.0.csv", sep=";")
metaData$Timepoint_dpi <- metaData$Timepoint..dpi.

# create pathogen column
metaData$pathogen <- ifelse(metaData$Experiment == "Heterodera schachtii 2", "Heterodera schachtii", metaData$Experiment)
metaData$pathogen <- ifelse(metaData$Experiment == "Hyaloperonospora arabidopsidis 3", "Hyaloperonospora arabidopsidis", metaData$pathogen)
metaData$pathogen <- ifelse(metaData$Experiment == "Hyaloperonospora arabidopsidis 2", "Hyaloperonospora arabidopsidis", metaData$pathogen)
metaData$pathogen <- ifelse(metaData$Experiment == "Plasmodiophora brassicae 2", "Plasmodiophora brassicae", metaData$pathogen)
metaData$pathogen <- ifelse(metaData$Experiment == "Turnip mosaic virus (TuMV) 3", "Turnip mosaic virus (TuMV)", metaData$pathogen)

# get pathogen abbreviations
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Heterodera schachtii", "H. schachtii", metaData$pathogen)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Alternaria brassicae", "A. brassicae", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Botrytis cinerea", "B. cinerea", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Cauliflower mosaic virus (CaMV)", "CaMV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Turnip mosaic virus (TuMV)", "TuMV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Turnip yellows virus (TuYV)", "TuYV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Turnip yellows virus (TuYV)", "TuYV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Golovinomyces orontii", "G. orontii", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Erysiphe cichoracearum ", "E. cichoracearum", metaData$pathogen_abbr) 	
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Hyaloperonospora arabidopsidis", "Hpa", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Meloidogyne javanica", "M. javanica", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Plasmodiophora brassicae", "P. brassicae", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Erisyphe cruciferarum", "E. cruciferarum", metaData$pathogen_abbr)

# column with infected pathogens (take all control samples together)
metaData$inf_path <- ifelse(metaData$infected_control == "infected", metaData$pathogen_abbr, "control")

# coumn with infected pathogens in category
metaData$Category <- ifelse(metaData$infected_control == "infected", metaData$Category, "control")
metaData$Category_trophs <- ifelse(metaData$infected_control == "infected", metaData$Category_trophs, "control")

# get library size in the metadata
kallisto_alignment_results <- read.csv('~/Studium/bachelor-thesis-luisa/multiqc/kallisto_alignment_summary.csv')
metaData <- merge(metaData, kallisto_alignment_results, by.x = "Run", by.y = "sample_ID", all.x = TRUE)
colnames(metaData) <- ifelse(colnames(metaData) == "n_pseudoaligned", "LibrarySize", colnames(metaData))
metaData <- metaData[order(metaData$ID),]
metaData$LibrarySize_kallisto <- metaData$LibrarySize


# create binary meta data for correlation calculations
metaData_bin <- binarizeCategoricalColumns(metaData, convertColumns = c("inf_path", "infected_control", "Timepoint_dpi", "Library", "Category", "Category_trophs", "Tissue"), includePairwise = FALSE, includeLevelVsAll = TRUE, minCount = 1, dropFirstLevelVsAll = FALSE)
colnames(metaData_bin) <- gsub(".vs.all", "", colnames(metaData_bin))

# remove CaMV category (unsure)
metaData_bin <- metaData_bin[,-50]
```

### Filter low expressed genes (with tpm values (genelevel tpm)): 

Result: 
- before filtering: 27598 genes
- after filtering: 25175 genes

if tiniest dataset with contrast is 3 samples, then
--> keep genes that have at least 0.5 TPM per sample in at least 3 samples.

```{r}
smallest_tpm <- 0.5
percentage <- 0.51
num_samples <- ncol(tpm_counts)
# genes to keep
keep <- rowSums(tpm_counts >= smallest_tpm) >= 3

gene_counts_filtered <- gene_counts[keep,]
```

### Variance stabilization transformation

VST on filtered gene count table:

```{r}
dds <- DESeqDataSetFromMatrix(countData = round(gene_counts_filtered),
                              colData = metaData,
                              design = ~Experiment + infected_control)

dds <- estimateSizeFactors(dds)
#summary(sizeFactors(dds))

dds <- estimateDispersions(dds)
plotDispEsts(dds)
#dds_genelev_filtered <- DESeq(dds_genelev_filtered)

vsd <- varianceStabilizingTransformation(dds, blind = FALSE)
#vsd_genelev_filtered_data <- getVarianceStabilizedData(dds_genelev_filtered)
```
Filter datasets if there are outliers:

```{r}
# removing samples with bad alignment rate
#experiment_to_drop <- 'Hyaloperonospora arabidopsidis 2'
#to_drop <- meta_info[meta_info$Experiment == experiment_to_drop, ]$Run
hpa2 <- c("SRR22565219","SRR22565220","SRR22565221","SRR22565222","SRR22565223","SRR22565224","SRR22565225","SRR22565226","SRR22565227","SRR22565228","SRR22565229","SRR22565230","SRR22565231","SRR22565232","SRR22565233","SRR22565234","SRR22565235","SRR22565236")

# remove samples with library size < 1.000.000 (four hpa samples)
remove <- metaData$Run[metaData$LibrarySize < 1000000]

gene_counts <- gene_counts[, !names(gene_counts) %in% hpa2]
tpm_counts <- tpm_counts[, !names(tpm_counts) %in% hpa2]

metaData <- metaData[!metaData$Run %in% hpa2,]
metaData_bin <- metaData_bin[!metaData_bin$Run %in% hpa2,]
```

Re-Do VST:

```{r}
dds <- DESeqDataSetFromMatrix(countData = round(all_gene_counts_filtered),
                              colData = metaData,
                              design = ~Experiment + infected_control)

dds <- estimateSizeFactors(dds)
#summary(sizeFactors(dds))

dds <- estimateDispersions(dds)
plotDispEsts(dds)
#dds_genelev_filtered <- DESeq(dds_genelev_filtered)

vsd <- varianceStabilizingTransformation(dds, blind = FALSE)
#vsd_genelev_filtered_data <- getVarianceStabilizedData(dds_genelev_filtered)
```
```{r}
# include librarysize in the meta Data
# librarySize (total number of mapped reads per sample) as summed up counts of all genes for each sample
library_sizes <- colSums(gene_counts_filtered)
metaData$LibrarySize <- library_sizes[match(metaData$Run, names(library_sizes))]
metaData_bin$LibrarySize <- metaData$LibrarySize
```

### Prinicipal Component Analysis (PCA)

PCA function:

```{r pca function}
perform_pca <- function(count){
   #log_count_matrix <- log2(count + 1)
  # perform PCA
  pca_result <- prcomp(t(count))
  return (pca_result)
}

visualize_standard_pca <- function(data, pca_result, ellipse, title, PC3){
  if(ellipse == TRUE){
    if(PC3 == TRUE){
      ggplot(data, aes(x = PC3, y = PC4, color = Experiment)) +
      geom_point() +
      stat_ellipse(aes(color = Experiment), type = "norm") +
      xlab(paste0("PC3: ", round(100 * summary(pca_result)$importance[2, 3], 2), "% variance")) +
      ylab(paste0("PC4: ", round(100 * summary(pca_result)$importance[2, 4], 2), "% variance")) +
      ggtitle(title) +
      theme_minimal()
    }
    else{
    ggplot(data, aes(x = PC1, y = PC2, color = Experiment)) +
      geom_point() +
      stat_ellipse(aes(color = Experiment), type = "norm") +
      xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
      ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
      ggtitle(title) +
      theme_minimal()
      }
  }
  
  else{
    if(PC3 == TRUE){
      ggplot(data, aes(x = PC3, y = PC4, color = Experiment)) +
      geom_point() +
      xlab(paste0("PC3: ", round(100 * summary(pca_result)$importance[2, 3], 2), "% variance")) +
      ylab(paste0("PC4: ", round(100 * summary(pca_result)$importance[2, 4], 2), "% variance")) +
      ggtitle(title) +
      theme_minimal()
    }
    else{
      ggplot(data, aes(x = PC1, y = PC2, color = Experiment)) +
        geom_point() +
        xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
        ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
        ggtitle(title) +
        theme_minimal()
      }
  }
}
```

PCA on raw (and filtered) vst counts:  
  
```{r pca}
# vst counts filtered with tpm value
pca <- perform_pca(assay(vsd))
pca_data <- data.frame(Sample = rownames(pca$x),
                       PC1 = pca$x[, 1], 
                       PC2 = pca$x[, 2], 
                       PC3 = pca$x[, 3], 
                       PC4 = pca$x[, 4]) %>% merge(metaData, by.x = "Sample", by.y = "Run")

pca_data$infected_control <- as.character(pca_data$infected_control)
```


PCA on filtered counts

```{r pca}
# PC1 and PC2
p_brassicae_root_infected <- metaData[metaData$infection_type == "infected - root",]
visualize_standard_pca(pca_data,pca, TRUE,"PCA on filtered vst counts", FALSE)
ggplot(pca_data, aes(x = PC1, y = PC2, color = Experiment, shape=infected_control)) +
        geom_point() +
        
        #stat_ellipse(aes(color = Category), type = "norm")+
        theme_minimal() +
        labs(
          title = "PCA on filtered vst counts",
          x = paste0("PC1: ", round(100 * summary(pca)$importance[2, 1], 2), "% variance"),
          y = paste0("PC2: ", round(100 * summary(pca)$importance[2, 2], 2), "% variance"), 
          color = "Experiment", 
          shape = "infected/control"
        ) +
        theme(
          legend.position = "right",
          plot.title = element_text(hjust = 0.5, size = 16)
        )
      

# PC3 and PC4
visualize_standard_pca(pca_data,pca, FALSE,"PCA on filtered vst counts", TRUE)
```

PCA on further information, e.g. Library, Tissues:

```{r pca}
# number of aligned reads per sample (absolute counts) to include in the PCA
ggplot(pca_data, aes(x = PC1, y = PC2, color = LibrarySize, size = LibrarySize)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  #stat_ellipse(aes(color = Experiment), type = "norm") +# Color gradient for aligned reads
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "PCA with Library Size",
    x = paste0("PC1: ", round(100 * summary(pca)$importance[2, 1], 2), "% variance"),
    y = paste0("PC2: ", round(100 * summary(pca)$importance[2, 2], 2), "% variance"), 
    size = "Library Size",
    color = "Library Size"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )


#PC3 and PC4
ggplot(pca_data, aes(x = PC3, y = PC4, color = LibrarySize, size = LibrarySize)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  #stat_ellipse(aes(color = Experiment), type = "norm") +# Color gradient for aligned reads
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "PCA with Library Size",
    x = paste0("PC3: ", round(100 * summary(pca)$importance[2, 3], 2), "% variance"),
    y = paste0("PC4: ", round(100 * summary(pca)$importance[2, 4], 2), "% variance"), 
    size = "Library Size",
    color = "Library Size"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )

# include single and paired in the PCA
ggplot(pca_data, aes(x = PC1, y = PC2, color = Library)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA with Library",
    x = paste0("PC1: ", round(100 * summary(pca)$importance[2, 1], 2), "% variance"),
    y = paste0("PC2: ", round(100 * summary(pca)$importance[2, 2], 2), "% variance"), 
    color = "single-end/paired-end"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )

ggplot(pca_data, aes(x = PC3, y = PC4, color = Library)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA with Library",
    x = paste0("PC3: ", round(100 * summary(pca)$importance[2, 3], 2), "% variance"),
    y = paste0("PC4: ", round(100 * summary(pca)$importance[2, 4], 2), "% variance"), 
    color = "single-end/paired-end"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )

# PCA on tissue samples
ggplot(pca_data, aes(x = PC1, y = PC2, color = Tissue)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA with Tissues",
    x = paste0("PC1: ", round(100 * summary(pca)$importance[2, 1], 2), "% variance"),
    y = paste0("PC2: ", round(100 * summary(pca)$importance[2, 2], 2), "% variance"), 
    color = "single-end/paired-end"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )
```


```{r}
# load expression data
options(stringsAsFactors = FALSE)
#expr_data <- read.csv("~/Studium/bachelor-thesis-luisa/count_tables/vst_expression_data.csv", row.names = 1)

# transpose expression data for WGCNA
expr_data <- assay(vsd)
input_mat0 <- t(expr_data)
```

```{r}
gsg = goodSamplesGenes(input_mat0, verbose = 3)
gsg$allOK # TRUE
summary(gsg)
```

```{r}
# Identifying outlier samples
sampleTree0 <- hclust(dist(input_mat0), method = "average") #Clustering samples based on distance 


#Plotting the cluster dendrogram
options(repr.plot.width = 20, repr.plot.height = 10)
#Setting the graphical parameters
par(cex = 0.6);
par(mar = c(0,4,2,0))
plot(sampleTree0, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
cex.axis = 1.5, cex.main = 2)
abline(h=300, col ="red")
```

Three outliers were found that should be removed --> choose a cut height of 300 to keep the rest of the samples

```{r}
clust = cutreeStatic(sampleTree0, cutHeight = 300, minSize = 10)
table(clust)

# clust 1 contains the samples we want to keep.
keepSamples = (clust==1)
input_mat = input_mat0[keepSamples, ]
nGenes = ncol(input_mat)
nSamples = nrow(input_mat)

# adjust meta data to filtered input matrix
metaSamples = rownames(input_mat)
traitRows = match(metaSamples, metaData$Run)
metaData = metaData[traitRows, ]
metaData_bin = metaData_bin[traitRows, ]

# Identifying outlier samples
sampleTree <- hclust(dist(input_mat), method = "average") #Clustering samples based on distance 


#Plotting the cluster dendrogram
options(repr.plot.width = 20, repr.plot.height = 10)
#Setting the graphical parameters
par(cex = 0.6);
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
cex.axis = 1.5, cex.main = 2)
```


```{r}
# only numerical meta Data for sample dendrogram with meta data
metData <- metaData_bin[, -c(1,2,3)]

sampleTree2 <- hclust(dist(input_mat), method = "average")

metaColors_sub_cat <- numbers2colors(metData[, grep("Category\\.", colnames(metData))], signed = FALSE)
options(repr.plot.width = 20, repr.plot.height = 20,repr.res = 200)
plotDendroAndColors(sampleTree2, metaColors_sub_cat, groupLabels = names(metData[, grep("Category\\.", colnames(metData))])%>% gsub("Category\\.", "", .), main = "Sample dendrogram and metadata heatmap")

metaColors_timepoint <- numbers2colors(as.numeric(metaData[, "Timepoint..dpi."]), signed = FALSE)
options(repr.plot.width = 20, repr.plot.height = 20,repr.res = 200)
plotDendroAndColors(sampleTree2, metaColors_timepoint, groupLabels = "Timepoints post-infection", main = "Sample dendrogram and metadata heatmap")

metaColors_tissues <- numbers2colors(metaData_bin[, grep("Tissue", colnames(metaData_bin))], signed = FALSE)
options(repr.plot.width = 20, repr.plot.height = 20,repr.res = 200)
plotDendroAndColors(sampleTree2, metaColors_tissues, groupLabels = names(metaData_bin[, grep("Tissue\\.", colnames(metaData_bin))])%>% gsub("Tissue\\.", "", .), main = "Sample dendrogram and metadata heatmap")

metData_all <- metaData_bin[, -c(1,2,3,4,10,11,31:48,62,64:69)]
metData_all <- metData_all[, -c(2,3,15:19)]

metData_all <- metData_all %>% mutate_all(as.numeric)
metaColors_all <- numbers2colors(metData_all, signed = FALSE)
#par(mar = c(0,10,2,0))
#options(repr.plot.width = 25, repr.plot.height = 20,repr.res = 200)
plotDendroAndColors(sampleTree2, metaColors_all, groupLabels = names(metData_all) %>% sub("^[^.]*\\.", "", .),main = "Sample dendrogram and metadata heatmap")

```

```{r}
# save expression and meta data
write.csv(input_mat, "~/Studium/bachelor-thesis-luisa/count_tables/expr_data_transposed.csv")
write.csv(expr_data, "~/Studium/bachelor-thesis-luisa/count_tables/expr_data.csv")
write.csv(metaData, "~/Studium/bachelor-thesis-luisa/data/metaData_modified.csv")
write.csv(metaData_bin, "~/Studium/bachelor-thesis-luisa/data/metaData_bin.csv")
```



