---
title: "WGCNA"
output: html_document
date: "2025-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages
library(WGCNA)
```


#WGCNA - basic steps

### Data cleaning
1. load normalized expression Data and transform so that rows = samples and cols = genes
2. Identify outlier genes with WGCNA package -> goodSamplesGenes
3. Identify outlier samples with hierarchical clustering

```{r}
# load expression data
options(stringsAsFactors = FALSE)
expr_data <- read.csv("~/Studium/bachelor-thesis-luisa/count_tables/vst_expression_data.csv", row.names = 1)

# transpose it for WGCNA
input_mat <- t(expr_data)
```

```{r}
gsg = goodSamplesGenes(input_mat, verbose = 3)
gsg$allOK # TRUE
summary(gsg)
```

```{r}
# Identifying outlier samples
sampleTree <- hclust(dist(input_mat), method = "average") #Clustering samples based on distance 

#Setting the graphical parameters
par(cex = 0.6);
par(mar = c(0,4,2,0))

#Plotting the cluster dendrogram
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
cex.axis = 1.5, cex.main = 2)
```


### Network construction
Identify some sort of similarity measurement between pairs of genes -> represents the concordance of gene expression profiles across samples --> Pearson correlation coefficient
1. pick soft threshold and call adjacency function

```{r}
allowWGCNAThreads()

# other method: set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to = 20, by = 2))
sft <- pickSoftThreshold(input_mat, powerVector = powers, verbose = 5)
```

```{r}
sizeGrWindow(9,5)
par(mfrow = c(1,2));
cex1 = 0.9;
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2", type = "n",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red");

abline(h = 0.9, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")
```


look at grey modules, number of modules, connectivity

Run WGCNA_netconstr.R on the cluster.
Results for different soft powers:

```{r}
# network construction with soft power 4
load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow6.RData')
moduleColors_pow6 <- mergedColors
MEs_pow6 <- mergedMEs
geneTree_pow6 <- geneTree

load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow7.RData')
moduleColors_pow7 <- mergedColors
MEs_pow7 <- mergedMEs
geneTree_pow7 <- geneTree

load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow5.RData')
moduleColors_pow5 <- mergedColors
MEs_pow5 <- mergedMEs
geneTree_pow5 <- geneTree

load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow8.RData')
moduleColors_pow8 <- mergedColors
MEs_pow8 <- mergedMEs
geneTree_pow8 <- geneTree
```



### Module construction
1a. use hierarchical clustering to cluster the network into modules (transform adjacency matrix into measures of gene dissimilarity) OR
1b. use Topological Overlap Matrix (TOMsimilarity()) to define dissimilarity
2. dissimilarity/distance measures are then clustered using linkage hierarchical clustering and a dendrogram of genes
3. identify the Module Eigengene (standardized gene expression profile for a given module) using moduleEigengenes() function

```{r}
# Topological Overlap Matrix
#TOM.dissimilarity <- 1-TOMsimilarity(adjacency)
```

```{r}
# Hierarchical Clustering Analysis

# creating the dendrogram
#geneTree <- hclust(as.dist(TOM.dissimilarity), method = "average")
# plotting
sizeGrWindow(12,9)
plot(geneTree_pow6, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity", 
labels = FALSE, hang = 0.04)

# start with minimum module size 30 as recommended by the authors of WGCNA
#Modules <- cutreeDynamic(dendro = geneTree, distM = TOM.dissimilarity, deepSplit = 2, pamRespectsDendro = FALSE, minClusterSize = 30)
table(moduleColors_pow6) # 25 modules created
#ModuleColors <- labels2colors(Modules) # assigns each module number a color
#write.csv(table(ModuleColors), "ModuleColors_sp8.csv") # returns the counts for each color (aka the number of genes within each module)
```


```{r}
# plots the gene dendrogram with the module colors
plotDendroAndColors(geneTree_pow6, 
                    moduleColors_pow6,
                    "Module",
                    dendroLabels = FALSE, 
                    hang = 0.03,
                    addGuide = TRUE, 
                    guideHang = 0.05,
                    main = "Gene dendrogram and module colors")
```

```{r}
# Module Eigengene Identification
#MElist <- moduleEigengenes(input_mat, colors = ModuleColors)
#MEs <- MElist$eigengenes
head(MEs_pow6)
```


### Module merging
1. cluster modules based on pairwise eigengene correlations and merge modules that have similar expression profiles

```{r}
# eigengene dissimilarity
ME.dissimilarity = 1-cor(MElist$eigengenes, use="complete") #Calculate eigengene dissimilarity

# construct a cluster tree
METree = hclust(as.dist(ME.dissimilarity), method = "average") #Clustering eigengenes 
par(mar = c(0,4,2,0)) #seting margin sizes
par(cex = 0.6);#scaling the graphic
plot(METree)
abline(h=.25, col = "red") #a height of .25 corresponds to correlation of .75
```

```{r}
merge <- mergeCloseModules(input_mat, ModuleColors, cutHeight = 0.25)

# merged module colors
mergedColors <- merge$colors
# eigengenes of the new merged modules
mergedMEs <- merge$newMEs
```

```{r}
# plot dendrogram of original and merged module colors
plotDendroAndColors(geneTree, 
                    cbind(ModuleColors, mergedColors),
                    c("original Module", "merged Module"),
                    dendroLabels = FALSE, 
                    hang = 0.03,
                    addGuide = TRUE, 
                    guideHang = 0.05,
                    main = "Gene dendrogram and module colors for original and merged modules")
```

```{r}
TOMplot(TOM.dissimilarity, geneTree, mergedColors)
```

WGCNA in one step:

```{r}
picked_power = 9
temp_cor <- cor       
cor <- WGCNA::cor         # Force it to use WGCNA cor function (fix a namespace conflict issue)
netwk <- blockwiseModules(input_mat,                # <= input here
                          
                          # == Adjacency Function ==
                          power = picked_power,                # <= power here
                          networkType = "unsigned",
                          
                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = 30,
                          maxBlockSize = 4000,
                          
                          # == Module Adjustments ==
                          reassignThreshold = 0, 
                          mergeCutHeight = 0.25, 
                          
                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = T, 
                          saveTOMFileBase = "TOM",
                          
                          # == Output Options
                          numericLabels = T,
                          verbose = 3)

cor <- temp_cor
```

```{r}
# network construction with soft power 6
load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow6.RData')
moduleColors_pow6 <- mergedColors
MEs_pow6 <- mergedMEs
geneTree_pow6 <- geneTree

load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow7.RData')
moduleColors_pow7 <- mergedColors
MEs_pow7 <- mergedMEs
geneTree_pow7 <- geneTree

load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow5.RData')
moduleColors_pow5 <- mergedColors
MEs_pow5 <- mergedMEs
geneTree_pow5 <- geneTree

load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow8.RData')
moduleColors_pow8 <- mergedColors
MEs_pow8 <- mergedMEs
geneTree_pow8 <- geneTree
```


### External Trait Matching
1. relate the network to external traits (meta Info)
2. calculate the correlation of a trait of interest with previously identified module eigengenes --> gene significance

eigengene = summary profile for each module
--> correlate eigengenes with external traits and look for the most significant associations

```{r}
# Define numbers of genes and samples
nGenes = ncol(input_mat);
nSamples = nrow(input_mat);
picked_module_colors = moduleColors_pow6

#library(caret)
# create meta data table with pathogens as columns
dummyVarsObject <- dummyVars(~ pathogen_abbr, data = metaData)
metaData_num2 <- cbind(metaData, predict(dummyVarsObject, newdata = metaData))
metaData_num2 <- metaData_num2[,-c(9,10)]

# Recalculate MEs with color labels
MEs0 = moduleEigengenes(input_mat, picked_module_colors)$eigengenes
MEs = orderMEs(MEs0)

# rest meta info
moduleTraitCor = cor(MEs, metaData_num[,-1], use = "p"); # pearson correlation
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);

# calculate correlation based on pathogens
moduleTraitCor_path = cor(MEs, metaData_num2[, grep("pathogen_abbr", colnames(metaData_num2))], use = "p") 
moduleTraitPvalue_path = corPvalueStudent(moduleTraitCor, nSamples)
```

```{r}
sizeGrWindow(10,6)
textMatrix = paste(signif(moduleTraitCor, 2), "\n(",
                   signif(moduleTraitPvalue, 1), ")", sep = "")

dim(textMatrix) = dim(moduleTraitCor)
par(mar = c(6, 8.5, 3, 3))

# plot meta info
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = names(metaData_num[,-1]),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = greenWhiteRed(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(-1,1),
               main = paste("Module-trait relationships"))

# plot pathogens
textMatrix_path = paste(signif(moduleTraitCor_path, 2), "\n(",
                   signif(moduleTraitPvalue_path, 1), ")", sep = "")

dim(textMatrix_path) = dim(moduleTraitCor_path)
par(mar = c(6, 8.5, 3, 3))

labeledHeatmap(Matrix = moduleTraitCor_path,
               xLabels = colnames(metaData_num2[, grep("pathogen_abbr", colnames(metaData_num2))]) %>% gsub("pathogen_abbr", "", .),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = greenWhiteRed(50),
               textMatrix = textMatrix_path,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(-1,1),
               main = paste("Module-trait relationships with Pathogens"))
```

```{r}
module_order = names(MEs0) %>% gsub("ME","", .)
# add experiment to data
MEs0$treatment = row.names(MEs0)
MEs0 <- merge(MEs0,metaData, by.x = "treatment", by.y = "Run", all = TRUE)

# plot data
mME = MEs0 %>%
  pivot_longer(-c(treatment, Timepoint_dpi, infected.control.num, infection, infected_control, single_paired, avgSpotLen, pathogen, Experiment, pathogen_abbr)) %>%
  mutate(
    name = gsub("ME", "", name),
    name = factor(name, levels = module_order)
  )

mME %>% ggplot(., aes(x=pathogen, y=name, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limit = c(-1,1)) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title = "Module-trait Relationships", y = "Modules", fill="corr")
```

We quantify associations of individual genes with our trait of interest (weight) by defining Gene Significance GS as (the absolute value of) the correlation between the gene and the trait. For each module, we also define a quantitative measure of module membership MM as the correlation of the module eigengene and the gene expression profile. This allows us to quantify the similarity of all genes on the array to every module.

```{r}
# Define variable weight containing the weight column of datTrait
weight = as.data.frame(datTraits$weight_g);
names(weight) = "weight"
# names (colors) of the modules
modNames = substring(names(MEs), 3)
geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples))
```

```{r}
# identify hub genes
hubGenes = chooseTopHubInEachModule(input_mat, picked_module_colors)
print(hubGenes) 
```


### Target Gene Identification
1. use gene significance along with the genes intramodular connectivity to identify potential target genes associated with a particular trait of interest

### Network Visualization of Eigengenes
1. study the relationship among found modules
2. quantify module similarity by calculating the pairwise correlation of representative eigengenes
```{r}
dissTOM = 1-TOMsimilarityFromExpr(input_mat, power = 6);
# Transform dissTOM with a power to make moderately strong connections more visible in the heatmap
plotTOM = dissTOM^7;
# Set diagonal to NA for a nicer plot
diag(plotTOM) = NA;
# Call the plot function
sizeGrWindow(9,9)
TOMplot(plotTOM, geneTree_pow6, moduleColors_pow6, main = "Network heatmap plot, all genes")
```


### Network analysis with functional annotation and gene ontology
--> export a list of gene identifiers that can be used as input for Mercator OR
Enrichment analysis directly within R:

-> organism-specific package: Genome wide annotaation for Arabidopsis: org.At.tair.db

```{r}
#library(clusterProfiler)
#library(org.At.tair.db)

genes_in_module <- colnames(input_mat)[moduleColors_pow6=="brown"]
# remove .1 from gene IDs
genes_in_module <- gsub("\\.1$", "", genes_in_module)
enriched_go <- enrichGO(gene = genes_in_module, OrgDb = org.At.tair.db, keyType = "TAIR", 
                        ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 0.05)

barplot(enriched_go, showCategory = 10)

simple_enriched_go <- simplify(enriched_go)

simple_enriched_go %>% filter(p.adjust < 0.03) %>%
ggplot(showCategory = 20,
  aes(GeneRatio, forcats::fct_reorder(Description, GeneRatio))) + 
  geom_segment(aes(xend=0, yend = Description)) +
  geom_point(aes(color=p.adjust, size = Count)) +
  scale_color_viridis_c(guide=guide_colorbar(reverse=TRUE)) +
  scale_size_continuous(range=c(1, 7)) +
  theme_minimal() + 
  xlab("Gene Ratio") +
  ylab(NULL) + 
  ggtitle("GO Enrichment of up-regulated genes")


```


