---
title: "Data cleaning & preprocessing"
output: html_document
date: "2025-01-27"
---
```{r}
library(WGCNA)
library(ggplot2)
library(DESeq2)
library(dplyr)
library(tidyr)
```

Kallisto alignment result: 

```{r}
kallisto_alignment_summary <- read.csv('~/Studium/bachelor-thesis-luisa/scripts/kallisto_alignment_summary.csv')
#poor_alignment_threshold <- 2500000
#kallisto_alignment_summary$alignment_status <- ifelse(kallisto_alignment_summary$n_pseudoaligned < poor_alignment_threshold, "Poor", "Good")

#bad_alignment <- kallisto_alignment_summary[kallisto_alignment_summary$alignment_status =="Poor",]

kallisto_alignment_summary$n_not_aligned <- kallisto_alignment_summary$n_processed - kallisto_alignment_summary$n_pseudoaligned

data_long <- kallisto_alignment_summary %>%
  select(Sample = sample_ID, Pseudoaligned = n_pseudoaligned, NotAligned = n_not_aligned) %>%
  pivot_longer(cols = c(Pseudoaligned, NotAligned), names_to = "AlignmentStatus", values_to = "Count")

ggplot(data_long, aes(x=Sample, y=Count, fill = AlignmentStatus)) + 
  geom_bar(stat="identity") +
  scale_fill_manual(values = c("Pseudoaligned" = "steelblue", "NotAligned" = "darkred")) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  scale_x_discrete(breaks = data_long$Sample[seq(1, nrow(data_long), by = 10)]) +  # Show every 10th sample
  
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) +
  labs(title = "Kallisto: Alignment Scores for 283 samples",
       x = "",
       y = "Number of aligned Reads",
       fill = "Alignment Status")

#data_long <- data_long[!data_long$Sample %in% to_drop,]
```

Loading count data (saved via tximport on cluster):

```{r count tables}
counts_geneID <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/counts_genelev_corrected.csv', row.names = 1)
duplicates <- c("ERR7738945.1", "ERR7738946.1", "ERR7738947.1")
counts_geneID <- counts_geneID[,!names(counts_geneID) %in% duplicates] # genelevels

counts_genelev_tpm <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/counts_tpm_abundance2.csv', row.names = 1)
counts_genelev_tpm <- counts_genelev_tpm[,!names(counts_genelev_tpm) %in% duplicates] # genelevel tpm

counts_lengthscaledTPM <- read.csv('~/Studium/bachelor-thesis-luisa/count_tables/counts_lengthscaledTPM.csv', row.names = 1) # counts scaled with lenghtscaledTPM
```


Loading meta data:

```{r}
metaData <- read.csv("~/Studium/bachelor-thesis-luisa/data/metaData.csv", sep=";")
metaData$Timepoint_dpi <- metaData$Timepoint..dpi.

# create pathogen column
metaData$pathogen <- ifelse(metaData$Experiment == "Heterodera schachtii 2", "Heterodera schachtii", metaData$Experiment)
metaData$pathogen <- ifelse(metaData$Experiment == "Hyaloperonospora arabidopsidis 3", "Hyaloperonospora arabidopsidis", metaData$pathogen)
metaData$pathogen <- ifelse(metaData$Experiment == "Plasmodiophora brassicae 2", "Plasmodiophora brassicae", metaData$pathogen)
metaData$pathogen <- ifelse(metaData$Experiment == "Turnip mosaic virus (TuMV) 3", "Turnip mosaic virus (TuMV)", metaData$pathogen)

# get pathogen abbreviations
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Heterodera schachtii", "H. schachtii", metaData$pathogen)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Alternaria brassicae", "A. brassicae", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Botrytis cinerea", "B. cinerea", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Cauliflower mosaic virus (CaMV)", "CaMV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Turnip mosaic virus (TuMV)", "TuMV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Turnip yellows virus (TuYV)", "TuYV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Turnip yellows virus (TuYV)", "TuYV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Golovinomyces orontii", "G. orontii", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Erysiphe cichoracearum ", "E. cichoracearum", metaData$pathogen_abbr) 	
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Hyaloperonospora arabidopsidis", "Hpa", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Meloidogyne javanica", "M. javanica", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Plasmodiophora brassicae", "P. brassicae", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Erisyphe cruciferarum", "E. cruciferarum", metaData$pathogen_abbr)

# column with infected pathogens (take all control samples together)
metaData$inf_path <- ifelse(metaData$infected_control == "infected", metaData$pathogen_abbr, "control")

# coumn with infected pathogens in category
metaData$Category <- ifelse(metaData$infected_control == "infected", metaData$Category, "control")
metaData$Category_trophs <- ifelse(metaData$infected_control == "infected", metaData$Category_trophs, "control")

# get library size in the metadata
kallisto_alignment_results <- read.csv('~/Studium/bachelor-thesis-luisa/multiqc/kallisto_alignment_summary.csv')
metaData <- merge(metaData, kallisto_alignment_results, by.x = "Run", by.y = "sample_ID", all.x = TRUE)
colnames(metaData) <- ifelse(colnames(metaData) == "n_pseudoaligned", "LibrarySize", colnames(metaData))
metaData <- metaData[order(metaData$ID),]

# create binary meta data for correlation calculations
metaData_bin <- binarizeCategoricalColumns(metaData, convertColumns = c("inf_path", "infected_control", "Timepoint_dpi", "Library", "Category", "Category_trophs"), includePairwise = FALSE, includeLevelVsAll = TRUE, minCount = 1, dropFirstLevelVsAll = FALSE)
colnames(metaData_bin) <- gsub(".vs.all", "", colnames(metaData_bin))

# remove CaMV category (unsure)
metaData_bin <- metaData_bin[,-35]
```


```{r meta info}
# removing samples with bad alignment rate
#experiment_to_drop <- 'Hyaloperonospora arabidopsidis 2'
#to_drop <- meta_info[meta_info$Experiment == experiment_to_drop, ]$Run
hpa2 <- c("SRR22565219","SRR22565220","SRR22565221","SRR22565222","SRR22565223","SRR22565224","SRR22565225","SRR22565226","SRR22565227","SRR22565228","SRR22565229","SRR22565230","SRR22565231","SRR22565232","SRR22565233","SRR22565234","SRR22565235","SRR22565236")

#meta_info <- meta_info[!meta_info$Run %in% to_drop,]
#meta_info_general <- meta_info_general[!meta_info_general$Run %in% to_drop,]
counts_geneID <- counts_geneID[, !names(counts_geneID) %in% hpa2]
counts_genelev_tpm <- counts_genelev_tpm[, !names(counts_genelev_tpm) %in% hpa2]
```

Filter low expressed genes (with tpm values (genelevel tpm)): 

Result: 
- before filtering: 27321 genes
- after filtering: 24991 genes

if tiniest dataset with contrast, is 3 samples, then
--> keep genes that have at least 0.5 TPM per sample in at least 3 samples.

```{r}
smallest_tpm <- 0.5
percentage <- 0.51
num_samples <- ncol(counts_genelev_tpm)
# genes to keep
keep <- rowSums(counts_genelev_tpm >= smallest_tpm) >= 3

counts_genelev_filtered <- counts_geneID[keep,]
```


VST on filtered gene count table:

```{r}
#counts_genelev_filtered <- counts_genelev_filtered[, !names(counts_genelev_filtered) %in% hpa2]
dds_genelev_filtered <- DESeqDataSetFromMatrix(countData = round(counts_genelev_filtered),
                              colData = metaData,
                              design = ~Experiment + infected.control)

dds_genelev_filtered <- estimateSizeFactors(dds_genelev_filtered)
summary(sizeFactors(dds_genelev_filtered))

dds_genelev_filtered <- estimateDispersions(dds_genelev_filtered)
plotDispEsts(dds)
#dds_genelev_filtered <- DESeq(dds_genelev_filtered)

vsd_genelev_filtered <- varianceStabilizingTransformation(dds_genelev_filtered, blind = FALSE)
#vsd_genelev_filtered_data <- getVarianceStabilizedData(dds_genelev_filtered)
```


PCA function:

```{r pca function}
perform_pca <- function(count){
   #log_count_matrix <- log2(count + 1)
  # perform PCA
  pca_result <- prcomp(t(count))
  return (pca_result)
}

visualize_standard_pca <- function(data, pca_result, ellipse, title, PC3){
  if(ellipse == TRUE){
    if(PC3 == TRUE){
      ggplot(data, aes(x = PC3, y = PC4, color = Experiment)) +
      geom_point() +
      stat_ellipse(aes(color = Experiment), type = "norm") +
      xlab(paste0("PC3: ", round(100 * summary(pca_result)$importance[2, 3], 2), "% variance")) +
      ylab(paste0("PC4: ", round(100 * summary(pca_result)$importance[2, 4], 2), "% variance")) +
      ggtitle(title) +
      theme_minimal()
    }
    else{
    ggplot(data, aes(x = PC1, y = PC2, color = Experiment)) +
      geom_point() +
      stat_ellipse(aes(color = Experiment), type = "norm") +
      xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
      ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
      ggtitle(title) +
      theme_minimal()
      }
  }
  
  else{
    if(PC3 == TRUE){
      ggplot(data, aes(x = PC3, y = PC4, color = Experiment)) +
      geom_point() +
      xlab(paste0("PC3: ", round(100 * summary(pca_result)$importance[2, 3], 2), "% variance")) +
      ylab(paste0("PC4: ", round(100 * summary(pca_result)$importance[2, 4], 2), "% variance")) +
      ggtitle(title) +
      theme_minimal()
    }
    else{
      ggplot(data, aes(x = PC1, y = PC2, color = Experiment)) +
        geom_point() +
        xlab(paste0("PC1: ", round(100 * summary(pca_result)$importance[2, 1], 2), "% variance")) +
        ylab(paste0("PC2: ", round(100 * summary(pca_result)$importance[2, 2], 2), "% variance")) +
        ggtitle(title) +
        theme_minimal()
      }
  }
}
```

PCA on raw (and filtered) vst counts:  
  
```{r pca}
# vst counts filtered with tpm value
pca_filtered <- perform_pca(assay(vsd_genelev_filtered))
pca_data_filtered <- data.frame(Sample = rownames(pca_filtered$x),
                       PC1 = pca_filtered$x[, 1], 
                       PC2 = pca_filtered$x[, 2], 
                       PC3 = pca_filtered$x[, 3], 
                       PC4 = pca_filtered$x[, 4]) %>% merge(metaData, by.x = "Sample", by.y = "Run")
```


PCA on unfiltered vs filtered counts

```{r pca}
# PC1 and PC2
visualize_standard_pca(pca_data_filtered,pca_filtered, FALSE,"PCA of filtered vst counts", FALSE)

# PC3 and PC4
visualize_standard_pca(pca_data_filtered,pca_filtered, FALSE,"PCA of filtered vst counts", TRUE)
```

PCA on further information:

```{r pca}
# number of aligned reads per sample (absolute counts) to include in the PCA
ggplot(pca_data_filtered, aes(x = PC1, y = PC2, color = n_pseudoaligned, size = n_pseudoaligned)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  #stat_ellipse(aes(color = Experiment), type = "norm") +# Color gradient for aligned reads
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "Filtered PCA Plot with Aligned Reads",
    x = "PC 1",
    y = "PC 2",
    size = "Aligned Reads",
    color = "Aligned Reads"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )


#PC3 and PC4
ggplot(pca_data_filtered, aes(x = PC3, y = PC4, color = n_pseudoaligned, size = n_pseudoaligned)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  #stat_ellipse(aes(color = Experiment), type = "norm") +# Color gradient for aligned reads
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "Filtered PCA Plot with Aligned Reads",
    x = "PC 3",
    y = "PC 4",
    size = "Aligned Reads",
    color = "Aligned Reads"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )

# include single and paired in the PCA
ggplot(pca_data_filtered, aes(x = PC1, y = PC2, color = single.paired)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA with filtered vst counts",
    x = "PC 1",
    y = "PC 2",
    color = "single-end/paired-end"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )

ggplot(pca_data_filtered, aes(x = PC3, y = PC4, color = single.paired)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA with filtered vst counts",
    x = "PC 3",
    y = "PC 4",
    color = "single-end/paired-end"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16)
  )
```


```{r}
# load expression data
options(stringsAsFactors = FALSE)
#expr_data <- read.csv("~/Studium/bachelor-thesis-luisa/count_tables/vst_expression_data.csv", row.names = 1)

# transpose expression data for WGCNA
expr_data <- vsd_genelev_filtered
input_mat <- t(expr_data)
```

```{r}
gsg = goodSamplesGenes(input_mat, verbose = 3)
gsg$allOK # TRUE
summary(gsg)
```

```{r}
# Identifying outlier samples
sampleTree <- hclust(dist(input_mat), method = "average") #Clustering samples based on distance 


#Plotting the cluster dendrogram
options(repr.plot.width = 20, repr.plot.height = 10)
#Setting the graphical parameters
par(cex = 0.6);
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
cex.axis = 1.5, cex.main = 2)
```


```{r}
# only numerical meta Data for sample dendrogram with meta data
metData <- metaData_bin[, -c(1,2,3)]

sampleTree2 <- hclust(dist(input_mat), method = "average")

metaColors <- numbers2colors(metaData_bin[, grep("inf_path", colnames(metaData_bin))], signed = FALSE)
options(repr.plot.width = 20, repr.plot.height = 20,repr.res = 200)
plotDendroAndColors(sampleTree2, metaColors, groupLabels = names(metaData_bin[, grep("inf_path", colnames(metaData_bin))])%>% gsub("inf_path.", "", .), main = "Sample dendrogram and metadata heatmap")

metaColors_sub_cat <- numbers2colors(metaData_bin[, grep("Category\\.", colnames(metaData_bin))], signed = FALSE)
options(repr.plot.width = 20, repr.plot.height = 20,repr.res = 200)
plotDendroAndColors(sampleTree2, metaColors_sub_cat, groupLabels = names(metaData_bin[, grep("Category\\.", colnames(metaData_bin))])%>% gsub("Category\\.", "", .), main = "Sample dendrogram and metadata heatmap")

```

```{r}
# save expression and meta data
write.csv(input_mat, "~/Studium/bachelor-thesis-luisa/count_tables/expr_data_transposed")
write.csv(expr_data, "~/Studium/bachelor-thesis-luisa/count_tables/expr_data")
write.csv(metaData, "~/Studium/bachelor-thesis-luisa/data/metaData_modified")
write.csv(metaData_bin, "~/Studium/bachelor-thesis-luisa/data/metaData_bin")
```



