---
title: "Weighted Gene Co-Expression Network Analysis"
output: github_document
date: "2025-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages
library(WGCNA)
library(dplyr)
library(ggplot2)
library(tidyr)
library(clusterProfiler)
library(org.At.tair.db)
library(caret)
```


#WGCNA - basic steps

### Data cleaning
1. load normalized expression Data and transform so that rows = samples and cols = genes
2. Identify outlier genes with WGCNA package -> goodSamplesGenes
3. Identify outlier samples with hierarchical clustering

```{r}
# load expression data
options(stringsAsFactors = FALSE)
expr_data <- read.csv("~/Studium/bachelor-thesis-luisa/count_tables/vst_expression_data.csv", row.names = 1)

# transpose it for WGCNA
input_mat <- t(expr_data)
```

```{r}
gsg = goodSamplesGenes(input_mat, verbose = 3)
gsg$allOK # TRUE
summary(gsg)
```

```{r}
# Identifying outlier samples
sampleTree <- hclust(dist(input_mat), method = "average") #Clustering samples based on distance 

#Setting the graphical parameters
par(cex = 0.6);
par(mar = c(0,4,2,0))

#Plotting the cluster dendrogram
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
cex.axis = 1.5, cex.main = 2)
```


### Network construction
Identify some sort of similarity measurement between pairs of genes -> represents the concordance of gene expression profiles across samples --> Pearson correlation coefficient
1. pick soft threshold and call adjacency function

```{r}
allowWGCNAThreads()

# set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to = 20, by = 2))
sft <- pickSoftThreshold(input_mat, powerVector = powers, verbose = 5)
```

```{r}
sizeGrWindow(9,5)
par(mfrow = c(1,2));
cex1 = 0.9;
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2", type = "n",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red");

abline(h = 0.9, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")
```

### Module construction
1a. use hierarchical clustering to cluster the network into modules (transform adjacency matrix into measures of gene dissimilarity) OR
1b. use Topological Overlap Matrix (TOMsimilarity()) to define dissimilarity
2. dissimilarity/distance measures are then clustered using linkage hierarchical clustering and a dendrogram of genes
3. identify the Module Eigengene (standardized gene expression profile for a given module) using moduleEigengenes() function

### Module merging
1. cluster modules based on pairwise eigengene correlations and merge modules that have similar expression profiles

Run WGCNA_netconstr.R or WGCNA_auto_netconstr.R on the cluster.

Results for different soft powers:

```{r}
# network construction with soft power 6
load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow6.RData')
moduleColors_pow6 <- mergedColors
MEs_pow6 <- mergedMEs
geneTree_pow6 <- geneTree

load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow7.RData')
moduleColors_pow7 <- mergedColors
MEs_pow7 <- mergedMEs
geneTree_pow7 <- geneTree

load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow5.RData')
moduleColors_pow5 <- mergedColors
MEs_pow5 <- mergedMEs
geneTree_pow5 <- geneTree

load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow8.RData')
moduleColors_pow8 <- mergedColors
MEs_pow8 <- mergedMEs
geneTree_pow8 <- geneTree
```

```{r}
# plots the gene dendrogram with the module colors
plotDendroAndColors(geneTree_pow6, 
                    moduleColors_pow6,
                    "Module",
                    dendroLabels = FALSE, 
                    hang = 0.03,
                    addGuide = TRUE, 
                    guideHang = 0.05,
                    main = "Gene dendrogram and module colors")
```

### External Trait Matching
1. relate the network to external traits (meta Info)
2. calculate the correlation of a trait of interest with previously identified module eigengenes --> gene significance

eigengene = summary profile for each module
--> correlate eigengenes with external traits and look for the most significant associations

```{r}
# loading meta data
metaData <- read.csv("~/Studium/bachelor-thesis-luisa/data/metaData.csv", sep=";")
metaData$Timepoint_dpi <- metaData$Timepoint..dpi.
metaData$infected_control <- metaData$infected.control.num
metaData$single_paired <- ifelse(metaData$single.paired =="single", 1,0)
# create pathogen column
metaData$pathogen <- ifelse(metaData$Experiment == "Heterodera schachtii 2", "Heterodera schachtii", metaData$Experiment)
metaData$pathogen <- ifelse(metaData$Experiment == "Hyaloperonospora arabidopsidis 3", "Hyaloperonospora arabidopsidis", metaData$pathogen)
metaData$pathogen <- ifelse(metaData$Experiment == "Plasmodiophora brassicae 2", "Plasmodiophora brassicae", metaData$pathogen)
metaData$pathogen <- ifelse(metaData$Experiment == "Turnip mosaic virus (TuMV) 3", "Turnip mosaic virus (TuMV)", metaData$pathogen)
metaData <- metaData[, -c(3,4,5)]

# get pathogen abbreviations
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Heterodera schachtii", "H. schachtii", metaData$pathogen)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Alternaria brassicae", "A. brassicae", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Botrytis cinerea", "B. cinerea", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Cauliflower mosaic virus (CaMV)", "CaMV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Turnip mosaic virus (TuMV)", "TuMV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Turnip yellows virus (TuYV)", "TuYV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Turnip yellows virus (TuYV)", "TuYV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Golovinomyces orontii", "G. orontii", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Erysiphe cichoracearum ", "E. cichoracearum", metaData$pathogen_abbr) 	
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Hyaloperonospora arabidopsidis", "Hpa", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Meloidogyne javanica", "M. javanica", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Plasmodiophora brassicae", "P. brassicae", metaData$pathogen_abbr)

metaData_num <- metaData[,c(1,5,6,7,8)]

samples = rownames(input_mat)
metaRows = match(samples, metaData_num$Run)
metData = metaData_num[metaRows, -1]
rownames(metData) <- metaData_num[metaRows, 1]
```

```{r}
# Define numbers of genes and samples
nGenes = ncol(input_mat);
nSamples = nrow(input_mat);
picked_module_colors = moduleColors_pow6


# create meta data table with pathogens as columns
dummyVarsObject <- dummyVars(~ pathogen_abbr, data = metaData)
metaData_num2 <- cbind(metaData, predict(dummyVarsObject, newdata = metaData))
metaData_num2 <- metaData_num2[,-c(9,10)]

# Recalculate MEs with color labels
MEs0 = moduleEigengenes(input_mat, picked_module_colors)$eigengenes
MEs = orderMEs(MEs0)

# rest meta info
moduleTraitCor = cor(MEs, metaData_num[,-1], use = "p"); # pearson correlation
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);

# calculate correlation based on pathogens
moduleTraitCor_path = cor(MEs, metaData_num2[, grep("pathogen_abbr", colnames(metaData_num2))], use = "p") 
moduleTraitPvalue_path = corPvalueStudent(moduleTraitCor, nSamples)
```

```{r}
sizeGrWindow(10,6)
textMatrix = paste(signif(moduleTraitCor, 2), "\n(",
                   signif(moduleTraitPvalue, 1), ")", sep = "")

dim(textMatrix) = dim(moduleTraitCor)
par(mar = c(6, 8.5, 3, 3))

# plot meta info
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = names(metaData_num[,-1]),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = greenWhiteRed(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(-1,1),
               main = paste("Module-trait relationships"))

# plot pathogens
textMatrix_path = paste(signif(moduleTraitCor_path, 2), "\n(",
                   signif(moduleTraitPvalue_path, 1), ")", sep = "")

dim(textMatrix_path) = dim(moduleTraitCor_path)
#par(mar = c(6, 8.5, 3, 3))

labeledHeatmap(Matrix = moduleTraitCor_path,
               xLabels = colnames(metaData_num2[, grep("pathogen_abbr", colnames(metaData_num2))]) %>% gsub("pathogen_abbr", "", .),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = greenWhiteRed(50),
               textMatrix = textMatrix_path,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(-1,1),
               main = paste("Module-trait relationships with Pathogens"))
```

```{r}
module_order = names(MEs0) %>% gsub("ME","", .)
# add experiment to data
MEs0$treatment = row.names(MEs0)
MEs0 <- merge(MEs0,metaData, by.x = "treatment", by.y = "Run", all = TRUE)

# plot data
mME = MEs0 %>%
  pivot_longer(-c(treatment, Timepoint_dpi, infected.control.num, infection, infected_control, single_paired, avgSpotLen, pathogen, Experiment, pathogen_abbr)) %>%
  mutate(
    name = gsub("ME", "", name),
    name = factor(name, levels = module_order)
  )

mME %>% ggplot(., aes(x=pathogen, y=name, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limit = c(-1,1)) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title = "Module-trait Relationships", y = "Modules", fill="corr")
```

We quantify associations of individual genes with our trait of interest (weight) by defining Gene Significance GS as (the absolute value of) the correlation between the gene and the trait. For each module, we also define a quantitative measure of module membership MM as the correlation of the module eigengene and the gene expression profile. This allows us to quantify the similarity of all genes on the array to every module.


```{r}
# identify hub genes
hubGenes = chooseTopHubInEachModule(input_mat, picked_module_colors)
print(hubGenes)
```


### Target Gene Identification
1. use gene significance along with the genes intramodular connectivity to identify potential target genes associated with a particular trait of interest
--> The higher the mean gene significance in a module, the more significantly related the module is to the clinical trait of interest (see WGCNA paper)
--> Using the gene significance you can identify genes that have a high significance for weight. Using the module membership measures you can identify genes with high module membership in interesting modules.

Connectivity: how connected a speficic node is in the network (how many nodes have high correlation with that node). High connectivity indicates a hub gene (central to many nodes). Whole Network connectivity - a measure for how well the node is connected throughout the entire system Intramodular connectivity - a measure for how well the node is connected within its assigned module. Also an indicator for how well that node belongs to its module. This is also known as module membership.

```{r}
### Module Membership ###
picked_MEs <- MEs_pow6
modNames = substring(names(picked_MEs), 3) #extract module names

#Calculate the module membership and the associated p-values for every gene and module
geneModuleMembership = as.data.frame(cor(input_mat, picked_MEs, use = "p"))
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples)) # calculate student p-value for given correlations
names(geneModuleMembership) = paste("MM", modNames, sep="")
names(MMPvalue) = paste("p.MM", modNames, sep="")
```


```{r}
### Gene significance ###

# either define one variable containing the pathogen column
TuMV = as.data.frame(metaData_num2$pathogen_abbrTuMV)
names(TuMV) = "TuMV"
rownames(TuMV) = rownames(metData)

# calculate gene significance for every pathogen
path_df <- metaData_num2[, -c(1:8)]
rownames(path_df) <- metaData_num2[,1]
names(path_df) <- c("A.brassicae", "B.cinerea", "CaMV", "Drought", "E.cichoracearum", "G.orontii", "H.schachtii", "Hpa","M.javanica", "P.brassicae", "TuMV", "TuYV")

geneTraitSignificance = as.data.frame(cor(input_mat,path_df, use = "p"))
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance), nSamples))
names(geneTraitSignificance) = paste("GS.", names(path_df), sep="")
names(GSPvalue) = paste("p.GS.", names(path_df), sep="")

# mean GS across modules for specific trait
geneTraitSignificance$Module <- picked_module_colors
meanGS_per_module <- aggregate(. ~ Module, data = geneTraitSignificance, mean)
ggplot(meanGS_per_module, aes(Module, (GS.TuMV), fill = Module)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle=90)) + ggtitle("Mean Gene Significance across modules for specific pathogen")
barplot(
  height = meanGS_per_module$GS.TuMV, 
  names.arg = meanGS_per_module$Module, 
  col = meanGS_per_module$Module,  # Uses module names as colors if they are valid
  las = 2,  # Rotate x-axis labels for readability
  main = "Mean Gene Significance across Modules for TuMV",
  ylab = "Mean Gene Significance"
)
```


```{r}
# look at specific module that has high significance with choosen trait
module = "lightcyan"
GS_pathogen = "GS.TuMV"
column = match(module, modNames)
moduleGenes = picked_module_colors==module

# in modules related to the trait of interest, genes with high module membership often also have high gene significance
#pdf("~/Studium/bachelor-thesis-luisa/plots/GS_vs_MM_TuMV_magenta.pdf")
verboseScatterplot(abs(geneModuleMembership[moduleGenes,column]),
abs(geneTraitSignificance[moduleGenes,GS_pathogen]),
xlab = paste("Module Membership in", module, "module"),
ylab = "Gene significance for TuMV",
main = paste("Module membership vs. gene significance\n"),
cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2, col = "cyan")
#dev.off()
```


```{r}
MET = orderMEs(cbind(picked_MEs, path_df))
# Plot the relationships among the eigengenes and the trait
par(cex = 0.9)
plotEigengeneNetworks(MET, "", marDendro = c(0,4,1,2), marHeatmap = c(5,4,1,2), cex.lab = 0.8, xLabelsAngle
= 90)

#sizeGrWindow(6,6);
par(cex = 1.0)
plotEigengeneNetworks(MET, "Eigengene dendrogram", marDendro = c(0,4,2,0),
plotHeatmaps = FALSE)
# Plot the heatmap matrix (note: this plot will overwrite the dendrogram plot)
par(cex = 1.0)
plotEigengeneNetworks(MET, "Eigengene adjacency heatmap",  marHeatmap = c(3,4,2,2),
plotDendrograms = FALSE, xLabelsAngle = 90)
```



### Network Visualization of Eigengenes
1. study the relationship among found modules
2. quantify module similarity by calculating the pairwise correlation of representative eigengenes




### Network analysis with functional annotation and gene ontology
--> export a list of gene identifiers that can be used as input for Mercator OR
Enrichment analysis directly within R:

-> organism-specific package: Genome wide annotation for Arabidopsis: org.At.tair.db


```{r}
# BP = biological processes
# CC = Cellular components
# MF = molecular functions

module_genes <- list()

# get genes from each module
for (module in modNames) {
  module_genes[[module]] <- colnames(input_mat)[moduleColors_pow6 == module] %>% gsub("\\.1$", "", .)
}

# get all GO enrichment results for each module
go_results <- list(BP = list(), CC = list(), MF = list())

for (module in names(module_genes)) {
  go_results$BP[[module]] <- enrichGO(
    gene = module_genes[[module]],
    OrgDb = org.At.tair.db,  
    keyType = "TAIR",
    ont = "BP",  
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
  
  go_results$CC[[module]] <- enrichGO(
    gene = module_genes[[module]],
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "CC",  
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
  
  go_results$MF[[module]] <- enrichGO(
    gene = module_genes[[module]],
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "MF", 
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
}


# Get KEGG pathway enrichment results
kegg_results <- list()

for (module in names(module_genes)) {
  kegg_results[[module]] <- enrichKEGG(
    gene = module_genes[[module]],
    organism = "ath", 
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
}


module_names <- unique(names(module_genes))
enrichment_table <- data.frame(Module = module_names, GO_BP = NA, GO_CC = NA, GO_MF = NA, KEGG = NA, stringsAsFactors = FALSE)


extract_top_term <- function(enrichment_result) {
  if (!is.null(enrichment_result) && nrow(enrichment_result@result) > 0) {
    return(enrichment_result@result$Description[1])  # return top enriched term
  } else {
    return(NA)  # NA if no enrichment
  }
}

for (module in module_names) {
  enrichment_table$GO_BP[enrichment_table$Module == module] <- extract_top_term(go_results$BP[[module]])
  enrichment_table$GO_CC[enrichment_table$Module == module] <- extract_top_term(go_results$CC[[module]])
  enrichment_table$GO_MF[enrichment_table$Module == module] <- extract_top_term(go_results$MF[[module]])
  enrichment_table$KEGG[enrichment_table$Module == module] <- extract_top_term(kegg_results[[module]])
}
print(enrichment_table)
#write.csv(enrichment_table, "~/Studium/bachelor-thesis-luisa/wgcna_results/module_enrichment.csv")
```


```{r}
# plotting of enrichment results of specific module
module = "brown"

# simplify enrichGO output by removing redundancy of enriched GO terms
# plot with gene ratio
simple_enriched_go <- simplify(go_results$BP[[module]])

simple_enriched_go %>% filter(p.adjust < 0.05) %>%
ggplot(showCategory = 20,
  aes(GeneRatio, forcats::fct_reorder(Description, GeneRatio))) + 
  geom_segment(aes(xend=0, yend = Description)) +
  geom_point(aes(color=p.adjust, size = Count)) +
  scale_color_viridis_c(guide=guide_colorbar(reverse=TRUE)) +
  scale_size_continuous(range=c(1, 7)) +
  theme_minimal() + 
  xlab("Gene Ratio") +
  ylab(NULL) + 
  ggtitle(paste("GO Enrichment of", module, "module"))

goplot(simple_enriched_go, showCategory = 5)
cnetplot(simple_enriched_go)
```

