---
title: "WGCNA"
output: html_document
date: "2025-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages
library(WGCNA)
```


#WGCNA - basic steps

### Data cleaning
1. load normalized expression Data and transform so that rows = samples and cols = genes
2. Identify outlier genes with WGCNA package -> goodSamplesGenes
3. Identify outlier samples with hierarchical clustering

```{r}
# load expression data
options(stringsAsFactors = FALSE)
expr_data <- read.csv("~/Studium/bachelor-thesis-luisa/count_tables/vst_expression_data.csv", row.names = 1)

# transpose it for WGCNA
input_mat <- t(expr_data)
```

```{r}
gsg = goodSamplesGenes(input_mat, verbose = 3)
gsg$allOK # TRUE
summary(gsg)
```

```{r}
# Identifying outlier samples
sampleTree <- hclust(dist(input_mat), method = "average") #Clustering samples based on distance 

#Setting the graphical parameters
par(cex = 0.6);
par(mar = c(0,4,2,0))

#Plotting the cluster dendrogram
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
cex.axis = 1.5, cex.main = 2)
```


### Network construction
Identify some sort of similarity measurement between pairs of genes -> represents the concordance of gene expression profiles across samples --> Pearson correlation coefficient
1. pick soft threshold and call adjacency function

```{r}
allowWGCNAThreads()

spt <- pickSoftThreshold(input_mat)
spt

# other method: set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to = 20, by = 2))
sft <- pickSoftThreshold(input_mat, powerVector = powers, verbose = 5)
```
```{r}
par(mar=c(1,1,1,1))
plot(spt$fitIndices[,1],spt$fitIndices[,2],
xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
main = paste("Scale independence"))
text(spt$fitIndices[,1],spt$fitIndices[,2],col="red")
abline(h=0.80,col="red")

par(mar=c(1,1,1,1))
plot(spt$fitIndices[,1], spt$fitIndices[,5],
xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
main = paste("Mean connectivity"))
text(spt$fitIndices[,1], spt$fitIndices[,5], labels= spt$fitIndices[,1],col="red")
```
```{r}
par(mfrow = c(1,2));
cex1 = 0.9;

plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red"
)
abline(h = 0.90, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")
```

pick threshold 4 ??

look at grey modules, number of modules, connectivity

```{r}
# adjacency with soft threshold 4
softPower <- 4
adjacency <- adjacency(input_mat, power = softPower)
```


### Module construction
1a. use hierarchical clustering to cluster the network into modules (transform adjacency matrix into measures of gene dissimilarity) OR
1b. use Topological Overlap Matrix (TOMsimilarity()) to define dissimilarity
2. dissimilarity/distance measures are then clustered using linkage hierarchical clustering and a dendrogram of genes
3. identify the Module Eigengene (standardized gene expression profile for a given module) using moduleEigengenes() function

```{r}
# Topological Overlap Matrix
TOM <- TOMsimilarity(adjacency)
TOM.dissimilarity <- 1-TOM
```

```{r}
# Hierarchical Clustering Analysis

# creating the dendrogram
geneTree <- hclust(as.dist(TOM.dissimilarity), method = "average")
# plotting
sizeGrWindow(12,9)
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity", 
labels = FALSE, hang = 0.04)

# start with minimum module size 30 as recommended by the authors of WGCNA
Modules <- cutreeDynamic(dendro = geneTree, distM = TOM.dissimilarity, deepSplit = 2, pamRespectsDendro = FALSE, minClusterSize = 30)
table(Modules) # 25 modules created
ModuleColors <- labels2colors(Modules) # assigns each module number a color
table(ModuleColors) # returns the counts for each color (aka the number of genes within each module)

# plots the gene dendrogram with the module colors
plotDendroAndColors(geneTree, 
                    ModuleColors,
                    "Module",
                    dendroLabels = FALSE, 
                    hang = 0.03,
                    addGuide = TRUE, 
                    guideHang = 0.05,
                    main = "Gene dendrogram and module colors")
```
```{r}
# Module Eigengene Identification
MElist <- moduleEigengenes(input_mat, colors = ModuleColors)
MEs <- MElist$eigengenes
head(MEs)
```


### Module merging
1. cluster modules based on pairwise eigengene correlations and merge modules that have similar expression profiles

```{r}
# eigengene dissimilarity
ME.dissimilarity = 1-cor(MElist$eigengenes, use="complete") #Calculate eigengene dissimilarity

# construct a cluster tree
METree = hclust(as.dist(ME.dissimilarity), method = "average") #Clustering eigengenes 
par(mar = c(0,4,2,0)) #seting margin sizes
par(cex = 0.6);#scaling the graphic
plot(METree)
abline(h=.25, col = "red") #a height of .25 corresponds to correlation of .75
```
```{r}
merge <- mergeCloseModules(input_mat, ModuleColors, cutHeight = 0.25)

# merged module colors
mergedColors <- merge$colors
# eigengenes of the new merged modules
mergedMEs <- merge$newMEs
```
```{r}
# plot dendrogram of original and merged module colors
plotDendroAndColors(geneTree, 
                    cbind(ModuleColors, mergedColors),
                    c("original Module", "merged Module"),
                    dendroLabels = FALSE, 
                    hang = 0.03,
                    addGuide = TRUE, 
                    guideHang = 0.05,
                    main = "Gene dendrogram and module colors for original and merged modules")
```

```{r}
#TOMplot(TOM.dissimilarity, geneTree, mergedColors)
```

### External Trait Matching
1. relate the network to external traits (meta Info)
2. calculate the correlation of a trait of interest with previously identified module eigengenes --> gene significance

### Target Gene Identification
1. use gene significance along with the genes intramodular connectivity to identify potential target genes associated with a particular trait of interest

### Network Visualization of Eigengenes
1. study the relationship among found modules
2. quantify module similarity by calculating the pairwise correlation of representative eigengenes

