---
title: "Weighted Gene Co-Expression Network Analysis"
output: github_document
date: "2025-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages
library(WGCNA)
library(dplyr)
library(ggplot2)
library(tidyr)
library(clusterProfiler)
library(org.At.tair.db)
library(caret)
library(tibble)
```



### Data cleaning
1. Load normalized expression Data and transform so that rows = samples and cols = genes
2. Identify outlier genes with WGCNA package -> goodSamplesGenes
3. Identify outlier samples with hierarchical clustering

```{r}
# load expression data
options(stringsAsFactors = FALSE)
#expr_data <- read.csv("~/Studium/bachelor-thesis-luisa/count_tables/vst_expression_data.csv", row.names = 1)

# transposed expression data for WGCNA
input_mat <- read.csv("~/Studium/bachelor-thesis-luisa/count_tables/expr_data_transposed.csv", row.names = 1)

# load meta data
metaData <- read.csv("~/Studium/bachelor-thesis-luisa/data/metaData_modified.csv")
metaData_bin <- read.csv("~/Studium/bachelor-thesis-luisa/data/metaData_bin.csv", sep =";")
```

```{r}
gsg = goodSamplesGenes(input_mat, verbose = 3)
gsg$allOK # TRUE
summary(gsg)
```

```{r}
sampleTree <- hclust(dist(input_mat), method = "average") # average linkage method

par(cex = 0.6);
par(mar = c(0,4,2,0))

plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
cex.axis = 1.5, cex.main = 2)
```


### Network construction


```{r}
allowWGCNAThreads()

# set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to = 20, by = 2))
sft <- pickSoftThreshold(input_mat, powerVector = powers, verbose = 5, networkType = "signed hybrid")
```

```{r}
sizeGrWindow(9,5)
par(mfrow = c(1,2));
cex1 = 0.9;
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2", type = "n",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red");

abline(h = 0.85, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")
```


Due to memory issues, run WGCNA_netconstr.R (step-by-step) or WGCNA_auto_netconstr.R (automatic) on the cluster for network construction and module detection.


Results for different soft powers:

```{r}
# network construction with soft power 6 - signed hybrid network
load('~/Studium/bachelor-thesis-luisa/wgcna_results/new/networkConstruction-pow6_signed.RData')
moduleColors_pow6_signed <- mergedColors
MEs_pow6_signed <- mergedMEs
geneTree_pow6_signed <- geneTree

# network construction with soft power 5 - signed hybrid network
load('~/Studium/bachelor-thesis-luisa/wgcna_results/new/networkConstruction-pow5_signed.RData')
moduleColors_pow5_signed <- mergedColors
MEs_pow5_signed <- mergedMEs
geneTree_pow5_signed <- geneTree

# network construction with soft power 4 - signed hybrid network
load('~/Studium/bachelor-thesis-luisa/wgcna_results/new/networkConstruction-pow4_signed.RData')
moduleColors_pow4_signed <- mergedColors
MEs_pow4_signed <- mergedMEs
geneTree_pow4_signed <- geneTree

# network construction with soft power 4 - signed hybrid network with TOMType signed
load('~/Studium/bachelor-thesis-luisa/wgcna_results/new/networkConstruction-pow4_TOMsigned.RData')
moduleColors_pow4_tomsigned <- mergedColors
MEs_pow4_tomsigned <- mergedMEs
geneTree_pow4_tomsigned <- geneTree

```

```{r}
# plots the gene dendrogram with the module colors
plotDendroAndColors(geneTree_pow4_signed, 
                    moduleColors_pow4_signed,
                    "Module",
                    dendroLabels = FALSE, 
                    hang = 0.03,
                    addGuide = TRUE, 
                    guideHang = 0.05,
                    main = "Gene dendrogram and module colors")
table(moduleColors_pow4_signed)
```

### External Trait Matching

```{r}
nGenes = ncol(input_mat);
nSamples = nrow(input_mat);
picked_module_colors = moduleColors_pow4_signed

MEs0 = moduleEigengenes(input_mat, picked_module_colors)$eigengenes
MEs = orderMEs(MEs0)
```


Discover Module trait relationships:

```{r}
# divide meta table into subsets

# infected pathogens
inf_path_df <- metaData_bin[, grep("inf_path", colnames(metaData_bin))]
colnames(inf_path_df) <- gsub("inf_path.", "", colnames(inf_path_df))

# pathogen categories
cat_df <- metaData_bin[, grep("Category_trophs", colnames(metaData_bin))]
colnames(cat_df) <- gsub("Category_trophs.", "", colnames(cat_df))
cat_df <- cat_df[,-1]

# pathogen subcategories
sub_cat_df <- metaData_bin[, grep("Category\\.", colnames(metaData_bin))]
colnames(sub_cat_df) <- gsub("Category\\.", "", colnames(sub_cat_df))

# librarySize
libsize_df <- data.frame(LibrarySize = metaData[, 17])

#latest_df <- metaData_bin[, grep("latest_infection_timepoint.", colnames(metaData_bin))]
#colnames(latest_df) <- gsub("latest_infection_timepoint.", "", colnames(latest_df))
#latest_df <- latest_df[,-1]
```


```{r}
moduleTraitCorrelation <- function(MEs, metaData, trait){
  moduleTraitCor = cor(MEs, metaData, use = "p"); # pearson correlation
  moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);
  
  sizeGrWindow(10,6)
  textMatrix = paste(signif(moduleTraitCor, 2), "\n(",
                     signif(moduleTraitPvalue, 1), ")", sep = "")
  
  dim(textMatrix) = dim(moduleTraitCor)
  par(mar = c(6, 13, 3, 2.2))
  
  labeledHeatmap(Matrix = moduleTraitCor,
                 xLabels = colnames(metaData),
                 yLabels = names(MEs),
                 xLabelsAngle = 30,
                 ySymbols = names(MEs),
                 colorLabels = TRUE,
                 colors = blueWhiteRed(50),
                 textMatrix = textMatrix,
                 setStdMargins = FALSE,
                 cex.text = 0.5,
                 cex.lab.x = 1,
                 cex.lab.y = 1,
                 zlim = c(-1,1),
                 legendLabel = "Pearson's correlation",
                 main = paste("Module-trait relationships -", trait))
}

moduleTraitCorrelation(MEs, inf_path_df[,-11], "Pathogens")
moduleTraitCorrelation(MEs, sub_cat_df, "Pathogens (subcategories)")
moduleTraitCorrelation(MEs, cat_df, "Pathogens (categories)")
moduleTraitCorrelation(MEs, libsize_df, "Library Size")
#moduleTraitCorrelation(MEs, latest_df, "Comparing latest infection timepoints")
```

Look for weak correlations and sum up the absolute correlation value:

```{r}
biotrophs <- c("E..cichoracearum", "E..cruciferarum", "G..orontii", "H..schachtii", "Hpa", "M..javanica", "P..brassicae", "TuMV", "TuYV")

module_trait_inf_path <- cor(MEs, inf_path_df, use = "p")

biotrophic_modules <- rowSums(abs(module_trait_inf_path[, biotrophs]))
ranked_modules <- sort(biotrophic_modules, decreasing = TRUE)
head(ranked_modules)
```


### Target Gene Identification

```{r}
modNames = substring(names(MEs), 3) #extract module names

# calculate the module membership and the associated p-values for every gene and module
geneModuleMembership = as.data.frame(cor(input_mat, MEs, use = "p"))
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples)) 
names(geneModuleMembership) = paste("MM", modNames, sep="")
names(MMPvalue) = paste("p.MM", modNames, sep="")
```


```{r}
geneTraitSignificance_inf_path = as.data.frame(cor(input_mat,inf_path_df, use = "p"))
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance_inf_path), nSamples))

geneTraitSignificance_cat = as.data.frame(cor(input_mat,cat_df, use = "p"))
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance_cat), nSamples))

geneTraitSignificance_sub_cat = as.data.frame(cor(input_mat,sub_cat_df, use = "p"))
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance_sub_cat), nSamples))

geneTraitSignificance_inf_path$Module <- picked_module_colors
geneTraitSignificance_cat$Module <- picked_module_colors
geneTraitSignificance_sub_cat$Module <- picked_module_colors

```


```{r}
meanGS_per_module_inf_path <- aggregate(. ~ Module, data = geneTraitSignificance_inf_path, mean)
meanGS_per_module_cat <- aggregate(. ~ Module, data = geneTraitSignificance_cat, mean)
meanGS_per_module_sub_cat <- aggregate(. ~ Module, data = geneTraitSignificance_sub_cat, mean)

par(mar = c(7, 6, 2.5, 2.2))
barplot(
    height = meanGS_per_module_sub_cat$`oomycete.biotroph`, 
    names.arg = meanGS_per_module_sub_cat$Module, 
    col = meanGS_per_module_sub_cat$Module,  # module name as colors
    las = 2,  
    main = "Mean Gene Significance across Modules for oomycete biotroph",
    ylab = "Mean Gene Significance"
)
```

Different ways on calculating hub genes:

intramodular connectivity (kIM): measures how connected a gene is within its module (intramodularConnectivity)

kME: module membership (simpler to calculate, has an associated p-value and can be compared across modules)

signedKME: Calculation of (signed) eigengene-based connectivity, also known as module membership, correlation of the gene with the corresponding module eigengene.

```{r}
# identify top hub gene per module
hubGenes = chooseTopHubInEachModule(input_mat, picked_module_colors)
print(hubGenes)

kWithin = softConnectivity(input_mat, type = "signed hybrid")

# identify more hub genes per module based on gene significance
gene_info_GS <- data.frame(
  Gene = colnames(input_mat),
  Module = picked_module_colors,  
  kME = apply(geneModuleMembership, 1, max),
  GS = abs(geneTraitSignificance_inf_path[,-15]),
  kWithin = kWithin
)

# select top hub genes per module based on kME and kWithin
hub_genes_GS <- gene_info_GS %>%
  group_by(Module) %>%
  arrange(desc(kWithin), desc(kME)) %>%
  slice_head(n = 5) 

# not based on gene significance
gene_info <- data.frame(
  Gene = colnames(input_mat),
  Module = picked_module_colors,
  kME = apply(geneModuleMembership, 1, max),  
  kWithin = kWithin  
)
hub_genes <- gene_info %>%
  group_by(Module) %>%
  arrange(desc(kWithin), desc(kME)) #%>%
  #slice_head(n = 10) 

# choose hub genes based on module membership (signedKME)
kMEtable <- signedKME(input_mat, MEs)

# scale by max = TRUE --> make it comparable across modules
kIM_scaled <- intramodularConnectivity.fromExpr(input_mat, picked_module_colors, networkType = "signed hybrid", scaleByMax = TRUE, power = 4)
kIM <- intramodularConnectivity.fromExpr(input_mat, picked_module_colors, networkType = "signed hybrid", power = 4)
```



```{r}
# look at specific module that has high significance with choosen trait
module = "darkgreen"
category = "nematode"
column = match(module, modNames)
moduleGenes = picked_module_colors==module

# in modules related to the trait of interest, genes with high module membership often also have high gene significance

#pdf("~/Studium/bachelor-thesis-luisa/plots/GS_vs_MM_TuMV_magenta.pdf")
verboseScatterplot(geneModuleMembership[moduleGenes,column],
geneTraitSignificance_sub_cat[moduleGenes,category],
xlab = paste("Module Membership in", module, "module"),
ylab = paste("Gene significance for ", category),
main = paste("Module membership vs. gene significance\n"),
cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2, col = module)
#abline(h=0.5, v=0.8, col="red")
#dev.off()
```

```{r}
# calculate hub genes based on a MM and a GS cutoff
hub_genes_GS_MM <- function(MM, GS, module, trait, MM_cutoff, GS_cutoff){
  
  MM_table <- MM %>% rownames_to_column(var = "GeneID")
  GS_table <- GS %>% rownames_to_column(var = "GeneID")
  
  hub_data <- merge(data.frame(MM = MM_table[, module],GeneID = MM_table[, "GeneID"]),data.frame(GS = GS_table[,trait], GeneID = GS_table[, "GeneID"]), by = "GeneID")
  
  hub_genes <- hub_data %>% filter(MM > MM_cutoff & GS > GS_cutoff)
  
  return(hub_genes)
}

hub_genes_biotroph_darkred <- hub_genes_GS_MM(geneModuleMembership, geneTraitSignificance_cat, "MMdarkred", "biotroph", 0.8, 0.5)
hub_genes_biotroph_royalblue <- hub_genes_GS_MM(geneModuleMembership, geneTraitSignificance_cat, "MMroyalblue", "biotroph", 0.8, 0.5)
hub_genes_oomycete_royalblue <- hub_genes_GS_MM(geneModuleMembership, geneTraitSignificance_sub_cat, "MMroyalblue", "oomycete.biotroph", 0.8, 0.5)
hub_genes_nematode_brown <- hub_genes_GS_MM(geneModuleMembership, geneTraitSignificance_sub_cat, "MMbrown", "nematode", 0.8, 0.5)
hub_genes_nematode_darkgreen <- hub_genes_GS_MM(geneModuleMembership, geneTraitSignificance_sub_cat, "MMdarkgreen", "nematode", 0.8, 0.3)
hub_genes_hpa_darkolivegreen <- hub_genes_GS_MM(geneModuleMembership, geneTraitSignificance_inf_path, "MMdarkolivegreen", "Hpa", 0.8, 0.4)
```


```{r}
MET = orderMEs(cbind(MEs, inf_path_df))
par(cex = 0.9)
#bottom, left, top, right
plotEigengeneNetworks(MET, "Eigengene dendrogram and adjacency heatmap - including pathogens", marDendro = c(1,5,2,2), marHeatmap = c(7,7,2,2), cex.lab = 0.8, xLabelsAngle = 90)

MET_cat <- orderMEs(cbind(MEs, cat_df))
par(cex = 0.9)
#bottom, left, top, right
plotEigengeneNetworks(MET_cat, "Eigengene dendrogram and adjacency heatmap - categories", marDendro = c(1,5,2,2), marHeatmap = c(7,7,2,2), cex.lab = 0.8, xLabelsAngle = 90)


MET_sub_cat <- orderMEs(cbind(MEs, sub_cat_df))
par(cex = 0.9)
#bottom, left, top, right
plotEigengeneNetworks(MET_sub_cat, "Eigengene dendrogram and adjacency heatmap - subcategories", marDendro = c(1,5,2,2), marHeatmap = c(7,7,2,2), cex.lab = 0.8, xLabelsAngle = 90)

dist_matrix <- dist(t(MEs))

# Perform hierarchical clustering
hclust_modules <- hclust(dist_matrix, method = "average")
# Plot dendrogram
#plot(hclust_modules, main = "Hierarchical Clustering of Module Eigengenes", labels = names(MEs), cex = 0.7)
plotEigengeneNetworks(orderMEs(MEs), "Eigengene dendrogram and adjacency heatmap", marDendro = c(1,5,2,2), marHeatmap = c(7,7,2,2), cex.lab = 0.8, xLabelsAngle = 90)
```


### Network analysis with functional annotation and gene ontology
--> export a list of gene identifiers that can be used as input for Mercator
Enrichment analysis directly within R:
-> organism-specific package: Genome wide annotation for Arabidopsis: org.At.tair.db


##### Functional enrichment for all genes per module:

```{r}
# BP = biological processes
# CC = Cellular components
# MF = molecular functions

module_genes <- list()

# get genes from each module
for (module in modNames) {
  module_genes[[module]] <- colnames(input_mat)[moduleColors_pow4_tomsigned == module]
}

# get all GO enrichment results for each module
go_results <- list(BP = list(), CC = list(), MF = list())

for (module in names(module_genes)) {
  go_results$BP[[module]] <- enrichGO(
    gene = module_genes[[module]],
    OrgDb = org.At.tair.db,  
    keyType = "TAIR",
    ont = "BP",  
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
  
  go_results$CC[[module]] <- enrichGO(
    gene = module_genes[[module]],
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "CC",  
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
  
  go_results$MF[[module]] <- enrichGO(
    gene = module_genes[[module]],
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "MF", 
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
}


# KEGG pathway enrichment results
kegg_results <- list()

for (module in names(module_genes)) {
  kegg_results[[module]] <- enrichKEGG(
    gene = module_genes[[module]],
    organism = "ath", 
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
}


module_names <- unique(names(module_genes))
enrichment_table <- data.frame(Module = module_names, GO_BP = NA, GO_CC = NA, GO_MF = NA, KEGG = NA, stringsAsFactors = FALSE)


extract_top_term <- function(enrichment_result) {
  if (!is.null(enrichment_result) && nrow(enrichment_result@result) > 0) {
    return(enrichment_result@result$Description[1])  # return top enriched term
  } else {
    return(NA)  # NA if no enrichment
  }
}

for (module in module_names) {
  enrichment_table$GO_BP[enrichment_table$Module == module] <- extract_top_term(go_results$BP[[module]])
  enrichment_table$GO_CC[enrichment_table$Module == module] <- extract_top_term(go_results$CC[[module]])
  enrichment_table$GO_MF[enrichment_table$Module == module] <- extract_top_term(go_results$MF[[module]])
  enrichment_table$KEGG[enrichment_table$Module == module] <- extract_top_term(kegg_results[[module]])
}
print(enrichment_table)
write.csv(enrichment_table, "~/Studium/bachelor-thesis-luisa/wgcna_results/module_enrichment.csv")
```


```{r}
# plotting of enrichment results of specific module
module = "brown"

# simplify enrichGO output by removing redundancy of enriched GO terms
simple_enriched_go <- simplify(go_results$BP[[module]])

simple_enriched_go %>% filter(p.adjust < 0.05) %>%
ggplot(showCategory = 20,
  aes(GeneRatio, forcats::fct_reorder(Description, GeneRatio))) + 
  geom_segment(aes(xend=0, yend = Description)) +
  geom_point(aes(color=p.adjust, size = Count)) +
  scale_color_viridis_c(guide=guide_colorbar(reverse=TRUE)) +
  scale_size_continuous(range=c(1, 7)) +
  theme_minimal() + 
  xlab("Gene Ratio") +
  ylab(NULL) + 
  ggtitle(paste("GO Enrichment of", module, "module"))

goplot(simple_enriched_go, showCategory = 5)
cnetplot(simple_enriched_go)
```

Functional enrichment for hub genes based on signedkME > 0.8:

```{r}
hub_genes_per_module <- list()

for (mod in modNames) {
  hub_genes_per_module[[mod]] <- rownames(kMEtable)[which(kMEtable[, paste0("kME", mod)] > 0.8)] 
}


go_results_hub_compared <- compareCluster(geneCluster = hub_genes_per_module,
                             fun = "enrichGO",
                             OrgDb = org.At.tair.db,
                             keyType = "TAIR",
                             ont = "BP",
                             pAdjustMethod = "BH")


go_result_hub_simplified <- simplify(go_results_hub_compared)

go_result_hub_simplified %>% 
  filter(p.adjust < 0.05) %>%
  ggplot(aes(x = Cluster, y = forcats::fct_reorder(Description, GeneRatio))) + 
  geom_segment(aes(xend = Cluster, yend = Description)) +
  geom_point(aes(color = p.adjust, size = Count)) +
  scale_color_viridis_c(guide = guide_colorbar(reverse = TRUE)) +
  scale_size_continuous(range = c(1, 7)) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 8)) + 
  xlab("Module") +
  ylab(NULL) + 
  ggtitle("GO Enrichment of Module Hub Genes")
```



```{r}
# get all GO enrichment results for hub genes of each module
go_results_hub_genes <- list(BP = list(), CC = list(), MF = list())


for (module in names(hub_genes_per_module)) {
  go_results_hub_genes$BP[[module]] <- enrichGO(
    gene = hub_genes_per_module[[module]],
    OrgDb = org.At.tair.db,  
    keyType = "TAIR",
    ont = "BP",  
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
  
  go_results_hub_genes$CC[[module]] <- enrichGO(
    gene = hub_genes_per_module[[module]],
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "CC",  
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
  
  go_results_hub_genes$MF[[module]] <- enrichGO(
    gene = hub_genes_per_module[[module]],
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "MF", 
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
}


# Get KEGG pathway enrichment results
kegg_hub_results <- list()

for (module in names(hub_genes_per_module)) {
  kegg_hub_results[[module]] <- enrichKEGG(
    gene = hub_genes_per_module[[module]],
    organism = "ath", 
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
}

#browseKEGG()
```


```{r}
saveRDS(go_results_hub_genes, file = "~/Studium/bachelor-thesis-luisa/data/go_results_hub_genes.rds")
saveRDS(kegg_hub_results, file = "~/Studium/bachelor-thesis-luisa/data/kegg_hub_genes.rds")
go_results_hub_genes <-readRDS("~/Studium/bachelor-thesis-luisa/data/go_results_hub_genes.rds")
kegg_hub_results <- readRDS("~/Studium/bachelor-thesis-luisa/data/kegg_hub_genes.rds")
```

```{r}
# plotting of enrichment results of specific module
module = "orange"

# simplify enrichGO output by removing redundancy of enriched GO terms
# plot with gene ratio
simple_enriched_go_bp <- simplify(go_results_hub_genes$BP[[module]])
simple_enriched_go_mf <- simplify(go_results_hub_genes$MF[[module]])
simple_enriched_go_cc <- simplify(go_results_hub_genes$CC[[module]])

simple_enriched_go_bp %>% filter(p.adjust < 0.05) %>%
ggplot(showCategory = 20,
  aes(GeneRatio, forcats::fct_reorder(Description, GeneRatio))) + 
  geom_segment(aes(xend=0, yend = Description)) +
  geom_point(aes(color=p.adjust, size = Count)) +
  scale_color_viridis_c(guide=guide_colorbar(reverse=TRUE)) +
  scale_size_continuous(range=c(1, 7)) +
  theme_minimal() + 
  xlab("Gene Ratio") +
  ylab(NULL) + 
  ggtitle(paste("GO Enrichment of", module, "module"))

goplot(simple_enriched_go_bp, showCategory = 5)
cnetplot(simple_enriched_go_bp)

kegg_module <- kegg_hub_results[[module]]
head(kegg_module, n = 20)
```

Functional enrichment based on hub genes defined with a MM and GS cutoff:


```{r}
ego <- enrichGO(gene = hub_genes_nematode_brown$GeneID,
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "BP", 
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05)

simple_ego <- simplify(ego)
simple_ego %>% filter(p.adjust < 0.05) %>%
  ggplot(showCategory = 20,
  aes(GeneRatio, forcats::fct_reorder(Description, GeneRatio))) + 
  geom_segment(aes(xend=0, yend = Description)) +
  geom_point(aes(color=p.adjust, size = Count)) +
  scale_color_viridis_c(guide=guide_colorbar(reverse=TRUE)) +
  scale_size_continuous(range=c(1, 7)) +
  theme_minimal() + 
  xlab("Gene Ratio") +
  ylab(NULL) + 
  ggtitle(paste("GO Enrichment of hub genes of brown module"))

goplot(simple_ego, showCategory = 5)
cnetplot(simple_ego)
```


