---
title: "Weighted Gene Co-Expression Network Analysis"
output: github_document
date: "2025-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages
library(WGCNA)
library(dplyr)
library(ggplot2)
library(tidyr)
library(clusterProfiler)
library(org.At.tair.db)
library(caret)
```


#WGCNA - basic steps

### Data cleaning
1. load normalized expression Data and transform so that rows = samples and cols = genes
2. Identify outlier genes with WGCNA package -> goodSamplesGenes
3. Identify outlier samples with hierarchical clustering

```{r}
# load expression data
options(stringsAsFactors = FALSE)
#expr_data <- read.csv("~/Studium/bachelor-thesis-luisa/count_tables/vst_expression_data.csv", row.names = 1)
# transposed expression data for WGCNA
input_mat <- read.csv("~/Studium/bachelor-thesis-luisa/count_tables/expr_data_transposed", row.names = 1)

# load meta data
metaData <- read.csv("~/Studium/bachelor-thesis-luisa/data/metaData_modified")
metaData_bin <- read.csv("~/Studium/bachelor-thesis-luisa/data/metaData_bin")
```

```{r}
gsg = goodSamplesGenes(input_mat, verbose = 3)
gsg$allOK # TRUE
summary(gsg)
```

```{r}
# Identifying outlier samples
sampleTree <- hclust(dist(input_mat), method = "average") #Clustering samples based on distance 

#Setting the graphical parameters
par(cex = 0.6);
par(mar = c(0,4,2,0))

#Plotting the cluster dendrogram
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
cex.axis = 1.5, cex.main = 2)
```


### Network construction
Identify some sort of similarity measurement between pairs of genes -> represents the concordance of gene expression profiles across samples --> Pearson correlation coefficient
1. pick soft threshold and call adjacency function

```{r}
allowWGCNAThreads()

# set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to = 20, by = 2))
sft <- pickSoftThreshold(input_mat, powerVector = powers, verbose = 5, networkType = "signed hybrid")
```

```{r}
sizeGrWindow(9,5)
par(mfrow = c(1,2));
cex1 = 0.9;
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2", type = "n",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red");

abline(h = 0.9, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")
```

### Module construction
1a. use hierarchical clustering to cluster the network into modules (transform adjacency matrix into measures of gene dissimilarity) OR
1b. use Topological Overlap Matrix (TOMsimilarity()) to define dissimilarity
2. dissimilarity/distance measures are then clustered using linkage hierarchical clustering and a dendrogram of genes
3. identify the Module Eigengene (standardized gene expression profile for a given module) using moduleEigengenes() function

### Module merging
1. cluster modules based on pairwise eigengene correlations and merge modules that have similar expression profiles

Run WGCNA_netconstr.R or WGCNA_auto_netconstr.R on the cluster.

Results for different soft powers:

```{r}
# network construction with soft power 6
load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow6.RData')
moduleColors_pow6 <- mergedColors
MEs_pow6 <- mergedMEs
geneTree_pow6 <- geneTree

load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow7.RData')
moduleColors_pow7 <- mergedColors
MEs_pow7 <- mergedMEs
geneTree_pow7 <- geneTree

load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow5.RData')
moduleColors_pow5 <- mergedColors
MEs_pow5 <- mergedMEs
geneTree_pow5 <- geneTree

load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow8.RData')
moduleColors_pow8 <- mergedColors
MEs_pow8 <- mergedMEs
geneTree_pow8 <- geneTree

# network construction with soft power 6 - signed hybrid network
load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow6_signed.RData')
moduleColors_pow6_signed <- mergedColors
MEs_pow6_signed <- mergedMEs
geneTree_pow6_signed <- geneTree

# network construction with soft power 5 - signed hybrid network
load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow5_signed.RData')
moduleColors_pow5_signed <- mergedColors
MEs_pow5_signed <- mergedMEs
geneTree_pow5_signed <- geneTree

# network construction with soft power 4 - signed hybrid network
load('~/Studium/bachelor-thesis-luisa/wgcna_results/networkConstruction-pow4_signed.RData')
moduleColors_pow4_signed <- mergedColors
MEs_pow4_signed <- mergedMEs
geneTree_pow4_signed <- geneTree
```

```{r}
# plots the gene dendrogram with the module colors
plotDendroAndColors(geneTree_pow4_signed, 
                    moduleColors_pow4_signed,
                    "Module",
                    dendroLabels = FALSE, 
                    hang = 0.03,
                    addGuide = TRUE, 
                    guideHang = 0.05,
                    main = "Gene dendrogram and module colors")
```

### External Trait Matching
1. relate the network to external traits (meta Info)
2. calculate the correlation of a trait of interest with previously identified module eigengenes --> gene significance

eigengene = summary profile for each module
--> correlate eigengenes with external traits and look for the most significant associations

skip this part:

```{r}
# loading meta data
metaData <- read.csv("~/Studium/bachelor-thesis-luisa/data/metaData.csv", sep=";")
metaData$Timepoint_dpi <- metaData$Timepoint..dpi.

# create pathogen column
metaData$pathogen <- ifelse(metaData$Experiment == "Heterodera schachtii 2", "Heterodera schachtii", metaData$Experiment)
metaData$pathogen <- ifelse(metaData$Experiment == "Hyaloperonospora arabidopsidis 3", "Hyaloperonospora arabidopsidis", metaData$pathogen)
metaData$pathogen <- ifelse(metaData$Experiment == "Plasmodiophora brassicae 2", "Plasmodiophora brassicae", metaData$pathogen)
metaData$pathogen <- ifelse(metaData$Experiment == "Turnip mosaic virus (TuMV) 3", "Turnip mosaic virus (TuMV)", metaData$pathogen)

# get pathogen abbreviations
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Heterodera schachtii", "H. schachtii", metaData$pathogen)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Alternaria brassicae", "A. brassicae", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Botrytis cinerea", "B. cinerea", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Cauliflower mosaic virus (CaMV)", "CaMV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Turnip mosaic virus (TuMV)", "TuMV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Turnip yellows virus (TuYV)", "TuYV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Turnip yellows virus (TuYV)", "TuYV", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Golovinomyces orontii", "G. orontii", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Erysiphe cichoracearum ", "E. cichoracearum", metaData$pathogen_abbr) 	
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Hyaloperonospora arabidopsidis", "Hpa", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Meloidogyne javanica", "M. javanica", metaData$pathogen_abbr)
metaData$pathogen_abbr <- ifelse(metaData$pathogen == "Plasmodiophora brassicae", "P. brassicae", metaData$pathogen_abbr)

# column with infected pathogens (take all control samples together)
metaData$inf_path <- ifelse(metaData$infected_control == "infected", metaData$pathogen_abbr, "control")

# coumn with infected pathogens in category
metaData$Category <- ifelse(metaData$infected_control == "infected", metaData$Category, "control")
metaData$Category_trophs <- ifelse(metaData$infected_control == "infected", metaData$Category_trophs, "control")

# get library size in the metadata
kallisto_alignment_results <- read.csv('~/Studium/bachelor-thesis-luisa/multiqc/kallisto_alignment_summary.csv')
metaData <- merge(metaData, kallisto_alignment_results, by.x = "Run", by.y = "sample_ID", all.x = TRUE)
colnames(metaData) <- ifelse(colnames(metaData) == "n_pseudoaligned", "LibrarySize", colnames(metaData))
metaData <- metaData[order(metaData$ID),]

# create binary meta data for correlation calculations
metaData_bin <- binarizeCategoricalColumns(metaData, convertColumns = c("inf_path", "infected_control", "Timepoint_dpi", "Library", "Category", "Category_trophs"), includePairwise = FALSE, includeLevelVsAll = TRUE, minCount = 1, dropFirstLevelVsAll = FALSE)
colnames(metaData_bin) <- gsub(".vs.all", "", colnames(metaData_bin))

# remove CaMV category (unsure)
metaData_bin <- metaData_bin[,-35]
```


```{r}
# Define numbers of genes and samples
nGenes = ncol(input_mat);
nSamples = nrow(input_mat);
picked_module_colors = moduleColors_pow4_signed

# Recalculate MEs with color labels
MEs0 = moduleEigengenes(input_mat, picked_module_colors)$eigengenes
MEs = orderMEs(MEs0)
```

Discover Module trait relationships:

```{r}
# divide meta table into subsets

# infected pathogens
inf_path_df <- metaData_bin[, grep("inf_path", colnames(metaData_bin))]
colnames(inf_path_df) <- gsub("inf_path.", "", colnames(inf_path_df))

# pathogen categories
cat_df <- metaData_bin[, grep("Category_trophs", colnames(metaData_bin))]
colnames(cat_df) <- gsub("Category_trophs.", "", colnames(cat_df))
cat_df <- cat_df[,-1]
# pathogen subcategories
sub_cat_df <- metaData_bin[, grep("Category\\.", colnames(metaData_bin))]
colnames(sub_cat_df) <- gsub("Category\\.", "", colnames(sub_cat_df))

# librarySize
libsize_df <- data.frame(LibrarySize = metaData[, 16])
```


```{r}
moduleTraitCorrelation <- function(MEs, metaData, trait){
  moduleTraitCor = cor(MEs, metaData, use = "p"); # pearson correlation
  moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);
  
  sizeGrWindow(10,6)
  textMatrix = paste(signif(moduleTraitCor, 2), "\n(",
                     signif(moduleTraitPvalue, 1), ")", sep = "")
  
  dim(textMatrix) = dim(moduleTraitCor)
  par(mar = c(6, 13, 3, 2.2))
  
  labeledHeatmap(Matrix = moduleTraitCor,
                 xLabels = colnames(metaData),
                 yLabels = names(MEs),
                 xLabelsAngle = 30,
                 ySymbols = names(MEs),
                 colorLabels = TRUE,
                 colors = blueWhiteRed(50),
                 textMatrix = textMatrix,
                 setStdMargins = FALSE,
                 cex.text = 0.5,
                 cex.lab.x = 1,
                 cex.lab.y = 1,
                 zlim = c(-1,1),
                 legendLabel = "Pearson's correlation",
                 main = paste("Module-trait relationships -", trait))
}

moduleTraitCorrelation(MEs, inf_path_df, "Pathogens")
moduleTraitCorrelation(MEs, sub_cat_df, "Pathogens (subcategories)")
moduleTraitCorrelation(MEs, cat_df, "Pathogens (categories)")
moduleTraitCorrelation(MEs, libsize_df, "Library Size")
```


We quantify associations of individual genes with our trait of interest (weight) by defining Gene Significance GS as (the absolute value of) the correlation between the gene and the trait. For each module, we also define a quantitative measure of module membership MM as the correlation of the module eigengene and the gene expression profile. This allows us to quantify the similarity of all genes on the array to every module.



### Target Gene Identification
1. use gene significance along with the genes intramodular connectivity to identify potential target genes associated with a particular trait of interest
--> The higher the mean gene significance in a module, the more significantly related the module is to the clinical trait of interest (see WGCNA paper)
--> Using the gene significance you can identify genes that have a high significance for weight. Using the module membership measures you can identify genes with high module membership in interesting modules.

Connectivity: how connected a speficic node is in the network (how many nodes have high correlation with that node). High connectivity indicates a hub gene (central to many nodes). Whole Network connectivity - a measure for how well the node is connected throughout the entire system Intramodular connectivity - a measure for how well the node is connected within its assigned module. Also an indicator for how well that node belongs to its module. This is also known as module membership.

```{r}
### Module Membership ###
picked_MEs <- MEs_pow4_signed
modNames = substring(names(picked_MEs), 3) #extract module names

#Calculate the module membership and the associated p-values for every gene and module
geneModuleMembership = as.data.frame(cor(input_mat, picked_MEs, use = "p"))
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples)) # calculate student p-value for given correlations
names(geneModuleMembership) = paste("MM", modNames, sep="")
names(MMPvalue) = paste("p.MM", modNames, sep="")
```


```{r}
### Gene significance ###

# either define one variable containing the pathogen column, e.g.:
TuMV = as.data.frame(metaData_num2$pathogen_abbrTuMV)
names(TuMV) = "TuMV"
rownames(TuMV) = rownames(metData)

rownames(inf_path_df) <- metaData[,1]
rownames(sub_cat_df) <- metaData[,1]
rownames(cat_df) <- metaData[,1]
```


```{r}
geneTraitSignificance_inf_path = as.data.frame(cor(input_mat,inf_path_df, use = "p"))
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance_inf_path), nSamples))

geneTraitSignificance_cat = as.data.frame(cor(input_mat,cat_df, use = "p"))
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance_cat), nSamples))

geneTraitSignificance_sub_cat = as.data.frame(cor(input_mat,sub_cat_df, use = "p"))
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance_sub_cat), nSamples))

geneTraitSignificance_inf_path$Module <- picked_module_colors
geneTraitSignificance_cat$Module <- picked_module_colors
geneTraitSignificance_sub_cat$Module <- picked_module_colors

meanGS_per_module_inf_path <- aggregate(. ~ Module, data = geneTraitSignificance_inf_path, mean)
meanGS_per_module_cat <- aggregate(. ~ Module, data = geneTraitSignificance_cat, mean)
meanGS_per_module_sub_cat <- aggregate(. ~ Module, data = geneTraitSignificance_sub_cat, mean)
```


```{r}
barplot(
    height = meanGS_per_module_inf_path$`Hpa`, 
    names.arg = meanGS_per_module_sub_cat$Module, 
    col = meanGS_per_module_sub_cat$Module,  # module name as colors
    las = 2,  
    main = "Mean Gene Significance across Modules for Hpa",
    ylab = "Mean Gene Significance"
)
```
```{r}
# identify top hub gene per module
hubGenes = chooseTopHubInEachModule(input_mat, picked_module_colors)
print(hubGenes)

# kME: maximum module membership score
# kWithin: Intramodular connectivity, measures how connected a gene is within its module.
kWithin = softConnectivity(input_mat, type = "signed hybrid")

# identify more hub genes per module based on gene significance
gene_info_GS <- data.frame(
  Gene = colnames(input_mat),
  Module = picked_module_colors,  
  kME = apply(geneModuleMembership, 1, max),
  GS = abs(geneTraitSignificance_inf_path[,-14]), # gene trait significance for trait of interest (without module column)
  kWithin = kWithin
)

# Select top hub genes per module based on kME and kWithin
hub_genes_GS <- gene_info_GS %>%
  group_by(Module) %>%
  arrange(desc(kWithin), desc(kME)) %>%
  slice_head(n = 5)  # Select top 5 hub genes per module


# not based on gene significance
gene_info <- data.frame(
  Gene = colnames(input_mat),
  Module = picked_module_colors,  # Module assignment from WGCNA
  kME = apply(geneModuleMembership, 1, max),  # Max module membership per gene
  kWithin = kWithin  # Intramodular connectivity
)

# Select top hub genes per module based on kWithin and kME
hub_genes <- gene_info %>%
  group_by(Module) %>%
  arrange(desc(kWithin), desc(kME)) %>%
  slice_head(n = 10)  # Select top 5 hub genes per module
```

```{r}
plot_meanGS <- function(moduleColors, GS, trait){
  # mean GS across modules for specific trait
  GS$Module <- moduleColors
  meanGS_per_module <- aggregate(. ~ Module, data = GS, mean)
  
  barplot(
    height = meanGS_per_module$TuMV, 
    names.arg = meanGS_per_module$Module, 
    col = meanGS_per_module$Module,  # module name as colors
    las = 2,  
    main = paste("Mean Gene Significance across Modules for", trait),
    ylab = "Mean Gene Significance"
  )
}
```




```{r}
# look at specific module that has high significance with choosen trait
module = "darkgrey"
category = "Hpa"
column = match(module, modNames)
moduleGenes = picked_module_colors==module

# in modules related to the trait of interest, genes with high module membership often also have high gene significance

#pdf("~/Studium/bachelor-thesis-luisa/plots/GS_vs_MM_TuMV_magenta.pdf")
verboseScatterplot(abs(geneModuleMembership[moduleGenes,column]),
abs(geneTraitSignificance_inf_path[moduleGenes,category]),
xlab = paste("Module Membership in", module, "module"),
ylab = paste("Gene significance for ", category),
main = paste("Module membership vs. gene significance\n"),
cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2, col = module)
#dev.off()
```


```{r}
MET = orderMEs(cbind(picked_MEs, inf_path_df))
# Plot the relationships among the eigengenes and the trait
par(cex = 0.9)
#bottom, left, top, right
plotEigengeneNetworks(MET, "Eigengene dendrogram and adjacency heatmap - including pathogens", marDendro = c(1,5,2,2), marHeatmap = c(7,7,2,2), cex.lab = 0.8, xLabelsAngle = 90)

MET_cat <- orderMEs(cbind(picked_MEs, cat_df))
# Plot the relationships among the eigengenes and the trait
par(cex = 0.9)
#bottom, left, top, right
plotEigengeneNetworks(MET_cat, "Eigengene dendrogram and adjacency heatmap - categories", marDendro = c(1,5,2,2), marHeatmap = c(7,7,2,2), cex.lab = 0.8, xLabelsAngle = 90)

MET_sub_cat <- orderMEs(cbind(picked_MEs, sub_cat_df))
# Plot the relationships among the eigengenes and the trait
par(cex = 0.9)
#bottom, left, top, right
plotEigengeneNetworks(MET_sub_cat, "Eigengene dendrogram and adjacency heatmap - subcategories", marDendro = c(1,5,2,2), marHeatmap = c(7,7,2,2), cex.lab = 0.8, xLabelsAngle = 90)

```



### Network Visualization of Eigengenes
1. study the relationship among found modules
2. quantify module similarity by calculating the pairwise correlation of representative eigengenes




### Network analysis with functional annotation and gene ontology
--> export a list of gene identifiers that can be used as input for Mercator OR
Enrichment analysis directly within R:

-> organism-specific package: Genome wide annotation for Arabidopsis: org.At.tair.db


```{r}
# BP = biological processes
# CC = Cellular components
# MF = molecular functions

module_genes <- list()

# get genes from each module
for (module in modNames) {
  module_genes[[module]] <- colnames(input_mat)[moduleColors_pow4_signed == module] %>% gsub("\\.1$", "", .)
}

# get all GO enrichment results for each module
go_results <- list(BP = list(), CC = list(), MF = list())

for (module in names(module_genes)) {
  go_results$BP[[module]] <- enrichGO(
    gene = module_genes[[module]],
    OrgDb = org.At.tair.db,  
    keyType = "TAIR",
    ont = "BP",  
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
  
  go_results$CC[[module]] <- enrichGO(
    gene = module_genes[[module]],
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "CC",  
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
  
  go_results$MF[[module]] <- enrichGO(
    gene = module_genes[[module]],
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "MF", 
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
}


# Get KEGG pathway enrichment results
kegg_results <- list()

for (module in names(module_genes)) {
  kegg_results[[module]] <- enrichKEGG(
    gene = module_genes[[module]],
    organism = "ath", 
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
}


module_names <- unique(names(module_genes))
enrichment_table <- data.frame(Module = module_names, GO_BP = NA, GO_CC = NA, GO_MF = NA, KEGG = NA, stringsAsFactors = FALSE)


extract_top_term <- function(enrichment_result) {
  if (!is.null(enrichment_result) && nrow(enrichment_result@result) > 0) {
    return(enrichment_result@result$Description[1])  # return top enriched term
  } else {
    return(NA)  # NA if no enrichment
  }
}

for (module in module_names) {
  enrichment_table$GO_BP[enrichment_table$Module == module] <- extract_top_term(go_results$BP[[module]])
  enrichment_table$GO_CC[enrichment_table$Module == module] <- extract_top_term(go_results$CC[[module]])
  enrichment_table$GO_MF[enrichment_table$Module == module] <- extract_top_term(go_results$MF[[module]])
  enrichment_table$KEGG[enrichment_table$Module == module] <- extract_top_term(kegg_results[[module]])
}
print(enrichment_table)
write.csv(enrichment_table, "~/Studium/bachelor-thesis-luisa/wgcna_results/module_enrichment.csv")
```


```{r}
# plotting of enrichment results of specific module
module = "darkgrey"

# simplify enrichGO output by removing redundancy of enriched GO terms
# plot with gene ratio
simple_enriched_go <- simplify(go_results$BP[[module]])

simple_enriched_go %>% filter(p.adjust < 0.05) %>%
ggplot(showCategory = 20,
  aes(GeneRatio, forcats::fct_reorder(Description, GeneRatio))) + 
  geom_segment(aes(xend=0, yend = Description)) +
  geom_point(aes(color=p.adjust, size = Count)) +
  scale_color_viridis_c(guide=guide_colorbar(reverse=TRUE)) +
  scale_size_continuous(range=c(1, 7)) +
  theme_minimal() + 
  xlab("Gene Ratio") +
  ylab(NULL) + 
  ggtitle(paste("GO Enrichment of", module, "module"))

goplot(simple_enriched_go, showCategory = 5)
cnetplot(simple_enriched_go)
```
```{r}
# functional enrichment on hub genes per module
module_hub_genes <- list()
rownames(hub_genes) <- hub_genes$Gene
# get genes from each module
for (module in modNames) {
  module_hub_genes[[module]] <- rownames(hub_genes)[hub_genes$Module == module] %>% gsub("\\.1$", "", .)
}

# get all GO enrichment results for each module
go_results_hub_genes <- list(BP = list(), CC = list(), MF = list())

for (module in names(module_hub_genes)) {
  go_results_hub_genes$BP[[module]] <- enrichGO(
    gene = module_hub_genes[[module]],
    OrgDb = org.At.tair.db,  
    keyType = "TAIR",
    ont = "BP",  
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
  
  go_results_hub_genes$CC[[module]] <- enrichGO(
    gene = module_hub_genes[[module]],
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "CC",  
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
  
  go_results_hub_genes$MF[[module]] <- enrichGO(
    gene = module_hub_genes[[module]],
    OrgDb = org.At.tair.db,
    keyType = "TAIR",
    ont = "MF", 
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
}


# Get KEGG pathway enrichment results
kegg_hub_results <- list()

for (module in names(module_hub_genes)) {
  kegg_hub_results[[module]] <- enrichKEGG(
    gene = module_hub_genes[[module]],
    organism = "ath", 
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )
}
```

```{r}
# plotting of enrichment results of specific module
module = "darkgrey"

# simplify enrichGO output by removing redundancy of enriched GO terms
# plot with gene ratio
simple_enriched_go <- simplify(go_results_hub_genes$BP[[module]])

simple_enriched_go %>% filter(p.adjust < 0.05) %>%
ggplot(showCategory = 20,
  aes(GeneRatio, forcats::fct_reorder(Description, GeneRatio))) + 
  geom_segment(aes(xend=0, yend = Description)) +
  geom_point(aes(color=p.adjust, size = Count)) +
  scale_color_viridis_c(guide=guide_colorbar(reverse=TRUE)) +
  scale_size_continuous(range=c(1, 7)) +
  theme_minimal() + 
  xlab("Gene Ratio") +
  ylab(NULL) + 
  ggtitle(paste("GO Enrichment of", module, "module"))

goplot(simple_enriched_go, showCategory = 5)
cnetplot(simple_enriched_go)
```